---
title: "county_models"
format: html
editor: visual
---

## All Random Spatial County Models

```{r}

library(tidyverse)
library(ranger)

# ==========================================
# 1. Define the County Modeling Function
# ==========================================

run_county_model <- function(data_file, year_label, model_type) {
  
  output_file <- paste0("data/results/county", year_label, "_", model_type, "_results.csv")
  
  cat("\n==========================================\n")
  cat("RUNNING MODEL:", toupper(model_type), "| YEAR:", year_label, "\n")
  cat("Output ->", output_file, "\n")
  cat("==========================================\n")
  
  # A. Load the data
  clean_data <- read_csv(data_file, show_col_types = FALSE)
  
  # B. Prepare data and formula based on Model Type
  if (model_type == "m1") {
    # M1: Embeddings ONLY (vote_density kept in df for later analysis, but excluded from predictors)
    data_rf <- clean_data %>%
      select(
        county_fips, 
        vote_density, 
        TRUMPSHARE = rep_share, 
        ends_with("_mean")
      ) %>%
      na.omit()
    
    rf_formula <- TRUMPSHARE ~ . - county_fips - vote_density - fold_id
    n_features <- ncol(data_rf) - 3 # Subtract FIPS, density, share
    
  } else if (model_type == "m2") {
    # M2: Embeddings + Demographics + Baseline spatial
    data_rf <- clean_data %>%
      select(
        county_fips, 
        TRUMPSHARE = rep_share,
        area_km2, vote_density, 
        pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, 
        unemployment_rate, civilian_labor_force,
        ends_with("_mean")
      ) %>%
      na.omit()
    
    rf_formula <- TRUMPSHARE ~ . - county_fips - fold_id
    n_features <- ncol(data_rf) - 2 # Subtract FIPS, share
    
  } else if (model_type == "m3") {
    # M3: Baseline Spatial ONLY
    data_rf <- clean_data %>%
      select(
        county_fips, 
        TRUMPSHARE = rep_share,
        area_km2, vote_density
      ) %>%
      na.omit()
    
    rf_formula <- TRUMPSHARE ~ . - county_fips - fold_id
    n_features <- ncol(data_rf) - 2
    
  } else if (model_type == "m4") {
    # M4: Demographics + Baseline spatial (NO EMBEDDINGS)
    data_rf <- clean_data %>%
      select(
        county_fips, 
        TRUMPSHARE = rep_share,
        area_km2, vote_density, 
        pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, 
        unemployment_rate, civilian_labor_force
      ) %>%
      na.omit()
    
    rf_formula <- TRUMPSHARE ~ . - county_fips - fold_id
    n_features <- ncol(data_rf) - 2
  }
  
  # Calculate mtry (Standard regression rule: features / 3, minimum of 1)
  my_mtry <- max(1, floor(n_features / 3))
  
  cat("Data Prepared. Counties:", nrow(data_rf), "| Predictors:", n_features, "| mtry:", my_mtry, "\n")
  
  # C. Setup 5-Fold Cross-Validation
  set.seed(10)
  data_rf <- data_rf %>%
    mutate(fold_id = sample(rep(1:5, length.out = n())))
  
  all_predictions <- list()
  
  # D. The CV Loop
  for(k in 1:5) {
    
    train_df <- data_rf %>% filter(fold_id != k)
    test_df  <- data_rf %>% filter(fold_id == k)
    
    rf_model <- ranger(
      formula    = rf_formula, 
      data       = train_df,
      num.trees  = 500,
      mtry       = my_mtry,
      importance = "impurity",
      seed       = 42
    )
    
    # Predict and Store
    preds <- predict(rf_model, data = test_df)$predictions
    test_df$Predicted_Share <- preds
    all_predictions[[k]] <- test_df
    
    cat("  Fold", k, "complete.\n")
  }
  
  # E. Combine Folds and Calculate Errors
  full_results <- bind_rows(all_predictions) %>%
    mutate(
      Error = Predicted_Share - TRUMPSHARE,
      Abs_Error = abs(Error)
    )
  
  # F. Save Results
  write_csv(full_results, output_file)
  cat("-> Saved successfully!\n")
}

# ==========================================
# 2. Execute All Models (12 Total)
# ==========================================

# Define file paths
file_16 <- "data/prep/counties/master_county_emb_16.csv"
file_20 <- "data/prep/counties/master_county_emb_20.csv"
file_24 <- "data/prep/counties/master_county_emb_24.csv"

# We can use a nested loop to run all 12 combinations elegantly!
years <- list("16" = file_16, "20" = file_20, "24" = file_24)
models <- c("m1", "m2", "m3", "m4")

for (yr_label in names(years)) {
  for (mod in models) {
    run_county_model(data_file = years[[yr_label]], year_label = yr_label, model_type = mod)
  }
}

cat("\nAll 12 models successfully trained and saved!\n")

```

## All Leave Out State Models

```{r}
library(tidyverse)
library(ranger)

# ==========================================
# 1. Define the Spatial CV Modeling Function
# ==========================================

run_spatial_county_model <- function(data_file, year_label, model_type) {
  
  output_file <- paste0("data/results/county", year_label, "_", model_type, "_results.csv")
  
  cat("\n==========================================\n")
  cat("RUNNING SPATIAL MODEL:", toupper(model_type), "| YEAR:", year_label, "\n")
  cat("Output ->", output_file, "\n")
  cat("==========================================\n")
  
  # A. Load the data
  clean_data <- read_csv(data_file, show_col_types = FALSE)
  
  # B. Prepare data and formula based on Model Type
  # NOTE: We MUST select state_po to assign folds, but EXCLUDE it from the formula later!
  if (model_type == "m5") {
    # M5 (formerly M1): Embeddings ONLY
    data_rf <- clean_data %>%
      select(
        county_fips, 
        state_po,         # Needed for spatial CV
        vote_density,     # Kept for later analysis
        TRUMPSHARE = rep_share, 
        ends_with("_mean")
      ) %>%
      na.omit()
    
    rf_formula <- TRUMPSHARE ~ . - county_fips - state_po - vote_density - fold_id
    n_features <- ncol(data_rf) - 4 # Subtract FIPS, state, density, share
    
  } else if (model_type == "m6") {
    # M6 (formerly M2): Embeddings + Demographics + Baseline spatial
    data_rf <- clean_data %>%
      select(
        county_fips, 
        state_po,
        TRUMPSHARE = rep_share,
        area_km2, vote_density, 
        pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, 
        unemployment_rate, civilian_labor_force,
        ends_with("_mean")
      ) %>%
      na.omit()
    
    rf_formula <- TRUMPSHARE ~ . - county_fips - state_po - fold_id
    n_features <- ncol(data_rf) - 3 # Subtract FIPS, state, share
    
  } else if (model_type == "m7") {
    # M7 (formerly M3): Baseline Spatial ONLY
    data_rf <- clean_data %>%
      select(
        county_fips, 
        state_po,
        TRUMPSHARE = rep_share,
        area_km2, vote_density
      ) %>%
      na.omit()
    
    rf_formula <- TRUMPSHARE ~ . - county_fips - state_po - fold_id
    n_features <- ncol(data_rf) - 3
    
  } else if (model_type == "m8") {
    # M8 (formerly M4): Demographics + Baseline spatial (NO EMBEDDINGS)
    data_rf <- clean_data %>%
      select(
        county_fips, 
        state_po,
        TRUMPSHARE = rep_share,
        area_km2, vote_density, 
        pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, 
        unemployment_rate, civilian_labor_force
      ) %>%
      na.omit()
    
    rf_formula <- TRUMPSHARE ~ . - county_fips - state_po - fold_id
    n_features <- ncol(data_rf) - 3
  }
  
  # Calculate mtry (Standard regression rule: features / 3, minimum of 1)
  my_mtry <- max(1, floor(n_features / 3))
  
  # C. Setup 5-Fold SPATIAL Cross-Validation
  set.seed(10)
  
  # 1. Extract unique states (handles 50 states + DC if present)
  unique_states <- unique(data_rf$state_po)
  
  # 2. Shuffle states and assign to 5 folds
  state_folds <- tibble(
    state_po = sample(unique_states),
    fold_id = rep(1:5, length.out = length(unique_states))
  )
  
  # 3. Join folds back to the main dataframe
  data_rf <- data_rf %>%
    left_join(state_folds, by = "state_po")
  
  cat("Data Prepared. Counties:", nrow(data_rf), "| States:", length(unique_states), "\n")
  cat("Predictors:", n_features, "| mtry:", my_mtry, "\n")
  
  all_predictions <- list()
  
  # D. The CV Loop
  for(k in 1:5) {
    
    train_df <- data_rf %>% filter(fold_id != k)
    test_df  <- data_rf %>% filter(fold_id == k)
    
    rf_model <- ranger(
      formula    = rf_formula, 
      data       = train_df,
      num.trees  = 500,
      mtry       = my_mtry,
      importance = "impurity",
      seed       = 42
    )
    
    # Predict and Store
    preds <- predict(rf_model, data = test_df)$predictions
    test_df$Predicted_Share <- preds
    all_predictions[[k]] <- test_df
    
    # Optional: Print number of states tested in this fold
    states_tested <- length(unique(test_df$state_po))
    cat("  Fold", k, "complete. States tested:", states_tested, "\n")
  }
  
  # E. Combine Folds and Calculate Errors
  full_results <- bind_rows(all_predictions) %>%
    mutate(
      Error = Predicted_Share - TRUMPSHARE,
      Abs_Error = abs(Error)
    )
  
  # F. Save Results
  write_csv(full_results, output_file)
  cat("-> Saved successfully to", output_file, "\n")
}

# ==========================================
# 2. Execute All 12 Spatial Models
# ==========================================

# Define file paths
file_16 <- "data/prep/counties/master_county_emb_16.csv"
file_20 <- "data/prep/counties/master_county_emb_20.csv"
file_24 <- "data/prep/counties/master_county_emb_24.csv"

# Map years to files
years <- list("16" = file_16, "20" = file_20, "24" = file_24)

# Models M5 through M8
models <- c("m5", "m6", "m7", "m8")

# Loop through all 3 years and all 4 models
for (yr_label in names(years)) {
  for (mod in models) {
    run_spatial_county_model(data_file = years[[yr_label]], year_label = yr_label, model_type = mod)
  }
}

cat("\nAll 12 Spatial CV models successfully trained and saved!\n")
```

## All Temp alone models

```{r}
library(tidyverse)
library(ranger)

# ==========================================
# 1. Define the Temporal + Random Spatial Modeling Function
# ==========================================

run_temporal_model <- function(train_file, test_file, train_yr, test_yr, model_type) {
  
  output_file <- paste0("data/results/county", train_yr, "-", test_yr, "_", model_type, "_results.csv")
  
  cat("\n==========================================\n")
  cat("RUNNING TEMPORAL + RANDOM SPATIAL CV:", toupper(model_type), "\n")
  cat("Train:", train_yr, "-> Predict:", test_yr, "\n")
  cat("Output ->", output_file, "\n")
  cat("==========================================\n")
  
  # A. Helper function to select the correct columns
  prep_data <- function(filepath, mod) {
    df <- read_csv(filepath, show_col_types = FALSE)
    
    if (mod == "m9") {
      # M9: Embeddings ONLY
      df <- df %>% select(county_fips, vote_density, TRUMPSHARE = rep_share, ends_with("_mean"))
    } else if (mod == "m10") {
      # M10: Embeddings + Demographics + Baseline spatial
      df <- df %>% select(county_fips, TRUMPSHARE = rep_share, area_km2, vote_density, 
                          pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, 
                          unemployment_rate, civilian_labor_force, ends_with("_mean"))
    } else if (mod == "m11") {
      # M11: Baseline Spatial ONLY
      df <- df %>% select(county_fips, TRUMPSHARE = rep_share, area_km2, vote_density)
    } else if (mod == "m12") {
      # M12: Demographics + Baseline spatial (NO EMBEDDINGS)
      df <- df %>% select(county_fips, TRUMPSHARE = rep_share, area_km2, vote_density, 
                          pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, 
                          unemployment_rate, civilian_labor_force)
    }
    return(na.omit(df))
  }
  
  # B. Load and prep train/test datasets
  train_df <- prep_data(train_file, model_type)
  test_df  <- prep_data(test_file, model_type)
  
  # C. Ensure perfect 80/20 mapping across years by finding common counties
  common_fips <- intersect(train_df$county_fips, test_df$county_fips)
  
  set.seed(42) # Ensures identical folds across all models
  county_folds <- tibble(
    county_fips = sample(common_fips),
    fold_id = rep(1:5, length.out = length(common_fips))
  )
  
  # Join the folds back to both datasets (this also drops any unmatched counties automatically)
  train_df <- train_df %>% inner_join(county_folds, by = "county_fips")
  test_df  <- test_df %>% inner_join(county_folds, by = "county_fips")
  
  # D. Set up the Formula (subtracting fold_id)
  if (model_type == "m9") {
    rf_formula <- TRUMPSHARE ~ . - county_fips - vote_density - fold_id
    n_features <- ncol(train_df) - 4 # Subtract FIPS, density, share, fold_id
  } else {
    rf_formula <- TRUMPSHARE ~ . - county_fips - fold_id
    n_features <- ncol(train_df) - 3 # Subtract FIPS, share, fold_id
  }
  
  my_mtry <- max(1, floor(n_features / 3))
  
  cat("Data Prepared. Common Counties:", length(common_fips), "\n")
  cat("Predictors:", n_features, "| mtry:", my_mtry, "\n")
  
  all_predictions <- list()
  
  # E. The 5-Fold Temporal + Spatial CV Loop
  for(k in 1:5) {
    
    # Train on 80% in Year 1
    train_fold <- train_df %>% filter(fold_id != k)
    
    # Test on the remaining 20% in Year 2
    test_fold  <- test_df %>% filter(fold_id == k)
    
    rf_model <- ranger(
      formula    = rf_formula, 
      data       = train_fold,
      num.trees  = 500,
      mtry       = my_mtry,
      importance = "impurity",
      seed       = 42
    )
    
    # Predict and Store
    preds <- predict(rf_model, data = test_fold)$predictions
    test_fold$Predicted_Share <- preds
    all_predictions[[k]] <- test_fold
    
    cat("  Fold", k, "complete. Train counties:", nrow(train_fold), "-> Test counties:", nrow(test_fold), "\n")
  }
  
  # F. Combine Folds and Calculate Errors
  full_results <- bind_rows(all_predictions) %>%
    mutate(
      Error = Predicted_Share - TRUMPSHARE,
      Abs_Error = abs(Error)
    )
  
  # G. Save Results
  write_csv(full_results, output_file)
  cat("-> Saved successfully to", output_file, "\n")
}

# ==========================================
# 2. Execute All 8 Temporal Models
# ==========================================

# Define file paths
file_16 <- "data/prep/counties/master_county_emb_16.csv"
file_20 <- "data/prep/counties/master_county_emb_20.csv"
file_24 <- "data/prep/counties/master_county_emb_24.csv"

# Set up the pairs (Train 16 -> Test 20, Train 20 -> Test 24)
temporal_pairs <- list(
  list(train = file_16, test = file_20, train_yr = "16", test_yr = "20"),
  list(train = file_20, test = file_24, train_yr = "20", test_yr = "24")
)

models <- c("m9", "m10", "m11", "m12")

# Run the loops
for (pair in temporal_pairs) {
  for (mod in models) {
    run_temporal_model(
      train_file = pair$train, 
      test_file  = pair$test, 
      train_yr   = pair$train_yr, 
      test_yr    = pair$test_yr, 
      model_type = mod
    )
  }
}

cat("\nAll 8 Temporal out-of-time models successfully trained and saved!\n")
```

## All Cross-temp models

```{r}
library(tidyverse)
library(ranger)

# ==========================================
# 1. Define the Spatio-Temporal Modeling Function
# ==========================================

run_spatiotemporal_model <- function(train_file, test_file, train_yr, test_yr, model_type) {
  
  output_file <- paste0("data/results/county", train_yr, "-", test_yr, "_", model_type, "_results.csv")
  
  cat("\n==========================================\n")
  cat("RUNNING SPATIO-TEMPORAL MODEL:", toupper(model_type), "\n")
  cat("Train Year:", train_yr, "| Predict Year:", test_yr, "\n")
  cat("Output ->", output_file, "\n")
  cat("==========================================\n")
  
  # A. Helper function to select correct columns and keep 'state_po' for folds
  prep_data <- function(filepath, mod) {
    df <- read_csv(filepath, show_col_types = FALSE)
    
    if (mod == "m13") {
      # M13: Embeddings ONLY
      df <- df %>% select(county_fips, state_po, vote_density, TRUMPSHARE = rep_share, ends_with("_mean"))
    } else if (mod == "m14") {
      # M14: Embeddings + Demographics + Baseline spatial
      df <- df %>% select(county_fips, state_po, TRUMPSHARE = rep_share, area_km2, vote_density, 
                          pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, 
                          unemployment_rate, civilian_labor_force, ends_with("_mean"))
    } else if (mod == "m15") {
      # M15: Baseline Spatial ONLY
      df <- df %>% select(county_fips, state_po, TRUMPSHARE = rep_share, area_km2, vote_density)
    } else if (mod == "m16") {
      # M16: Demographics + Baseline spatial (NO EMBEDDINGS)
      df <- df %>% select(county_fips, state_po, TRUMPSHARE = rep_share, area_km2, vote_density, 
                          pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, 
                          unemployment_rate, civilian_labor_force)
    }
    return(na.omit(df))
  }
  
  # B. Load and prep train/test datasets
  train_df <- prep_data(train_file, model_type)
  test_df  <- prep_data(test_file, model_type)
  
  # C. Set up the Formula and mtry
  if (model_type == "m13") {
    rf_formula <- TRUMPSHARE ~ . - county_fips - state_po - vote_density - fold_id
    n_features <- ncol(train_df) - 4 # Subtract FIPS, state, density, share
  } else {
    rf_formula <- TRUMPSHARE ~ . - county_fips - state_po - fold_id
    n_features <- ncol(train_df) - 3 # Subtract FIPS, state, share
  }
  
  my_mtry <- max(1, floor(n_features / 3))
  
  # D. Setup 5-Fold SPATIAL Split (Using Train Year States)
  set.seed(10)
  
  unique_states <- unique(train_df$state_po)
  state_folds <- tibble(
    state_po = sample(unique_states),
    fold_id = rep(1:5, length.out = length(unique_states))
  )
  
  # Join folds back to BOTH train and test sets
  train_df <- train_df %>% left_join(state_folds, by = "state_po")
  test_df  <- test_df %>% left_join(state_folds, by = "state_po")
  
  cat("Data Prepared. Train Rows:", nrow(train_df), "| Test Rows:", nrow(test_df), "\n")
  cat("Predictors:", n_features, "| mtry:", my_mtry, "\n")
  
  all_predictions <- list()
  
  # E. The Spatio-Temporal CV Loop
  for(k in 1:5) {
    
    # TRAIN: 40 states in Year 1
    train_fold <- train_df %>% filter(fold_id != k)
    
    # TEST: 10 out-of-bag states in Year 2
    test_fold  <- test_df %>% filter(fold_id == k)
    
    rf_model <- ranger(
      formula    = rf_formula, 
      data       = train_fold,
      num.trees  = 500,
      mtry       = my_mtry,
      importance = "impurity",
      seed       = 42
    )
    
    # Predict on Year 2 states
    preds <- predict(rf_model, data = test_fold)$predictions
    test_fold$Predicted_Share <- preds
    all_predictions[[k]] <- test_fold
    
    states_trained <- length(unique(train_fold$state_po))
    states_tested  <- length(unique(test_fold$state_po))
    cat("  Fold", k, "complete. Trained on", states_trained, "states -> Tested on", states_tested, "states.\n")
  }
  
  # F. Combine Folds and Calculate Errors
  full_results <- bind_rows(all_predictions) %>%
    mutate(
      Error = Predicted_Share - TRUMPSHARE,
      Abs_Error = abs(Error)
    )
  
  # G. Save Results
  write_csv(full_results, output_file)
  cat("-> Saved successfully to", output_file, "\n")
}

# ==========================================
# 2. Execute All 8 Spatio-Temporal Models
# ==========================================

# Define file paths
file_16 <- "data/prep/counties/master_county_emb_16.csv"
file_20 <- "data/prep/counties/master_county_emb_20.csv"
file_24 <- "data/prep/counties/master_county_emb_24.csv"

# Set up the pairs (Train 16 -> Test 20, Train 20 -> Test 24)
temporal_pairs <- list(
  list(train = file_16, test = file_20, train_yr = "16", test_yr = "20"),
  list(train = file_20, test = file_24, train_yr = "20", test_yr = "24")
)

models <- c("m13", "m14", "m15", "m16")

# Run the loops
for (pair in temporal_pairs) {
  for (mod in models) {
    run_spatiotemporal_model(
      train_file = pair$train, 
      test_file  = pair$test, 
      train_yr   = pair$train_yr, 
      test_yr    = pair$test_yr, 
      model_type = mod
    )
  }
}

cat("\nAll 8 Spatio-Temporal models successfully trained and saved!\n")
```

What are the metrics for the in-year models?

```{r}
library(tidyverse)

# ==========================================
# Evaluate All Saved Models
# ==========================================

# Define the grid of years and models to loop through
years <- c("16", "20", "24")
models <- c("m1", "m2", "m3", "m4", "m5", "m6", "m7", "m8")

# Initialize an empty list to hold our metric rows
metrics_list <- list()

cat("Calculating metrics for all models...\n")

for (yr in years) {
  for (mod in models) {
    
    # Construct the exact file path we used to save them
    file_path <- paste0("data/results/county", yr, "_", mod, "_results.csv")
    
    # Safety check: only read the file if it actually exists
    if (file.exists(file_path)) {
      
      # Read the data quietly
      df <- read_csv(file_path, show_col_types = FALSE)
      
      # Calculate the core metrics
      # We use the 'Error' column you already built into the output (Predicted - Actual)
      mse  <- mean(df$Error^2, na.rm = TRUE)
      rmse <- sqrt(mse)
      
      rss  <- sum(df$Error^2, na.rm = TRUE)
      tss  <- sum((df$TRUMPSHARE - mean(df$TRUMPSHARE, na.rm = TRUE))^2, na.rm = TRUE)
      r2   <- 1 - (rss / tss)
      
      # Save this model's stats as a single-row tibble
      metrics_list[[paste0(yr, "_", mod)]] <- tibble(
        Year      = paste0("20", yr),
        Model     = toupper(mod),
        R_Squared = round(r2, 4),
        RMSE      = round(rmse, 4),
        MSE       = round(mse, 4)
      )
    } else {
      cat("Could not find file:", file_path, "\n")
    }
  }
}

# Combine the list into one beautiful master dataframe
all_model_metrics <- bind_rows(metrics_list) %>%
  arrange(Year, Model)

# ==========================================
# Print the Final Scoreboard
# ==========================================
cat("\n=======================================================\n")
cat("                 MODEL SCOREBOARD")
cat("\n=======================================================\n")
print(as.data.frame(all_model_metrics), row.names = FALSE)
cat("=======================================================\n")
```

What are the metrics for the cross-year models?

```{r}
library(tidyverse)

# ==========================================
# Evaluate Temporal Models
# ==========================================

# Define the train-test pairs and the models
temporal_pairs <- c("16-20", "20-24")
models <- c("m9", "m10", "m11", "m12", "m13", "m14", "m15", "m16")

# Initialize an empty list
metrics_list <- list()

cat("Calculating metrics for Temporal Models (M9-M12)...\n")

for (pair in temporal_pairs) {
  for (mod in models) {
    
    # Construct the exact file path for the temporal models
    file_path <- paste0("data/results/county", pair, "_", mod, "_results.csv")
    
    # Safety check: only read if it exists
    if (file.exists(file_path)) {
      
      df <- read_csv(file_path, show_col_types = FALSE)
      
      # Calculate the core metrics
      mse  <- mean(df$Error^2, na.rm = TRUE)
      rmse <- sqrt(mse)
      
      rss  <- sum(df$Error^2, na.rm = TRUE)
      tss  <- sum((df$TRUMPSHARE - mean(df$TRUMPSHARE, na.rm = TRUE))^2, na.rm = TRUE)
      r2   <- 1 - (rss / tss)
      
      # Save this model's stats
      metrics_list[[paste0(pair, "_", mod)]] <- tibble(
        Train_Test = pair,
        Model      = toupper(mod),
        R_Squared  = round(r2, 4),
        RMSE       = round(rmse, 4),
        MSE        = round(mse, 4)
      )
    } else {
      cat("Could not find file:", file_path, "\n")
    }
  }
}

# Combine into one master dataframe
temporal_metrics <- bind_rows(metrics_list) %>%
  arrange(Train_Test, Model)

# ==========================================
# Print the Final Scoreboard
# ==========================================
cat("\n=======================================================\n")
cat("           TEMPORAL MODEL SCOREBOARD (M9-M12)")
cat("\n=======================================================\n")
print(as.data.frame(temporal_metrics), row.names = FALSE)
cat("=======================================================\n")

```

## All change models

```{r}
library(tidyverse)
library(ranger)

# ==========================================
# 1. Load Data and Find Common Embeddings
# ==========================================

cat("Loading datasets...\n")
df_16 <- read_csv("data/prep/counties/clean_county_emb_16_17.csv", show_col_types = FALSE)
df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv", show_col_types = FALSE)
df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv", show_col_types = FALSE)

emb_cols_16 <- names(df_16) %>% keep(~ str_detect(.x, "_mean$"))
emb_cols_20 <- names(df_20) %>% keep(~ str_detect(.x, "_mean$"))
emb_cols_24 <- names(df_24) %>% keep(~ str_detect(.x, "_mean$"))
common_emb <- Reduce(intersect, list(emb_cols_16, emb_cols_20, emb_cols_24))

cat("Found", length(common_emb), "shared *_mean embedding columns.\n")

# ==========================================
# 2. Prepare 16-20 and 20-24 Datasets
# ==========================================

prep_shift_data <- function(df_early, df_late, early_suffix, late_suffix) {
  # Clean and rename base tables
  d1 <- df_early %>% select(county_fips, rep_share, total_votes, all_of(common_emb)) %>%
    rename_with(~ paste0(.x, "_", early_suffix), all_of(common_emb))
  
  d2 <- df_late %>% select(county_fips, county_name, state_po, vote_density, rep_share, total_votes, all_of(common_emb)) %>%
    rename_with(~ paste0(.x, "_", late_suffix), all_of(common_emb))
  
  # Join and calculate targets/turnout
  joined <- inner_join(d1, d2, by = "county_fips") %>%
    mutate(
      SHIFT = rep_share.y - rep_share.x,
      VOTE_CHANGE = (total_votes.y - total_votes.x) / total_votes.x
    ) %>%
    select(county_fips, county_name, state_po, vote_density, 
           rep_share_early = rep_share.x, rep_share_late = rep_share.y, 
           SHIFT, VOTE_CHANGE, everything(), -total_votes.x, -total_votes.y)
  
  # Calculate Deltas safely
  deltas <- purrr::map_dfc(common_emb, function(nm) {
    joined[[paste0(nm, "_", late_suffix)]] - joined[[paste0(nm, "_", early_suffix)]]
  })
  names(deltas) <- paste0(common_emb, "_delta")
  
  bind_cols(joined, deltas) %>% na.omit()
}

cat("\nPreparing specific election cycle datasets...\n")
data_1620 <- prep_shift_data(df_16, df_20, "16", "20")
data_2024 <- prep_shift_data(df_20, df_24, "20", "24")

# ==========================================
# 3. Apply 2x IQR Filter for Turnout Changes
# ==========================================

filter_turnout_iqr <- function(df, period_name) {
  n_before <- nrow(df)
  q1 <- quantile(df$VOTE_CHANGE, 0.25, na.rm = TRUE)
  q3 <- quantile(df$VOTE_CHANGE, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 2 * iqr
  upper_bound <- q3 + 2 * iqr
  
  df_clean <- df %>% filter(VOTE_CHANGE >= lower_bound & VOTE_CHANGE <= upper_bound)
  n_after <- nrow(df_clean)
  
  cat("[IQR Filter |", period_name, "] Kept", n_after, "of", n_before, 
      "counties. (Bounds:", round(lower_bound*100, 1), "% to", round(upper_bound*100, 1), "%)\n")
  
  return(df_clean %>% select(-VOTE_CHANGE))
}

data_1620 <- filter_turnout_iqr(data_1620, "2016-2020")
data_2024 <- filter_turnout_iqr(data_2024, "2020-2024")

# ==========================================
# 4. Define Helper to Select Predictors
# ==========================================

select_model_data <- function(df, model_num) {
  # All models keep these identifiers and the target
  base_cols <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_early", "rep_share_late", "SHIFT")
  
  if (model_num %in% c("m1", "m4")) {
    # M1/M4: Late year embeddings (e.g., 2020 for 16-20 shift)
    target_suffix <- if("A01_mean_20" %in% names(df)) "_20$" else "_24$"
    return(df %>% select(all_of(base_cols), matches(target_suffix)))
  } else if (model_num %in% c("m2", "m5")) {
    # M2/M5: Early year embeddings (e.g., 2016 for 16-20 shift)
    target_suffix <- if("A01_mean_16" %in% names(df)) "_16$" else "_20$"
    return(df %>% select(all_of(base_cols), matches(target_suffix)))
  } else if (model_num %in% c("m3", "m6")) {
    # M3/M6: Delta embeddings
    return(df %>% select(all_of(base_cols), ends_with("_delta")))
  }
}

# ==========================================
# 5. Modeling Function: Random Spatial CV
# ==========================================

run_spatial_change <- function(df, period_name, model_num) {
  out_name <- paste0("data/results/change", period_name, "_", model_num, "_results.csv")
  cat("\n--- Running Spatial", toupper(model_num), "for", period_name, "---\n")
  
  mod_df <- select_model_data(df, model_num)
  
  set.seed(42)
  mod_df <- mod_df %>% mutate(fold_id = sample(rep(1:5, length.out = n())))
  
  n_features <- ncol(mod_df) - 8 # Drop IDs + Target + Fold
  my_mtry <- min(8, max(1, floor(n_features / 3)))
  
  all_preds <- list()
  for (k in 1:5) {
    train_fold <- mod_df %>% filter(fold_id != k)
    test_fold  <- mod_df %>% filter(fold_id == k)
    
    rf <- ranger(
      SHIFT ~ . - county_fips - county_name - state_po - vote_density - rep_share_early - rep_share_late - fold_id,
      data = train_fold, num.trees = 500, mtry = my_mtry, importance = "impurity", seed = 42
    )
    
    test_fold$Predicted_Shift <- predict(rf, data = test_fold)$predictions
    all_preds[[k]] <- test_fold
  }
  
  results <- bind_rows(all_preds) %>%
    mutate(Error = Predicted_Shift - SHIFT, Abs_Error = abs(Error))
  
  write_csv(results, out_name)
  cat("Saved ->", out_name, "\n")
}

# ==========================================
# 6. Modeling Function: Temporal Transfer
# ==========================================

run_temporal_change <- function(train_df, test_df, model_num) {
  out_name <- paste0("data/results/change_", model_num, "_results.csv")
  cat("\n--- Running Temporal", toupper(model_num), "(Train: 16-20 -> Test: 20-24) ---\n")
  
  # Ensure column names match perfectly by removing the suffixes for training
  train_mod <- select_model_data(train_df, model_num)
  names(train_mod) <- str_replace(names(train_mod), "_16$|_20$|_24$", "")
  
  test_mod <- select_model_data(test_df, model_num)
  names(test_mod) <- str_replace(names(test_mod), "_16$|_20$|_24$", "")
  
  set.seed(55)
  train_mod <- train_mod %>% mutate(fold_id = sample(rep(1:5, length.out = n())))
  
  n_features <- ncol(train_mod) - 8 
  my_mtry <- min(8, max(1, floor(n_features / 3)))
  
  test_preds_by_fold <- list()
  
  for (k in 1:5) {
    t_fold <- train_mod %>% filter(fold_id != k)
    
    rf <- ranger(
      SHIFT ~ . - county_fips - county_name - state_po - vote_density - rep_share_early - rep_share_late - fold_id,
      data = t_fold, num.trees = 500, mtry = my_mtry, importance = "impurity", seed = 42
    )
    
    tmp <- test_mod %>% mutate(Predicted_Shift = predict(rf, data = test_mod)$predictions)
    test_preds_by_fold[[k]] <- tmp
  }
  
  # Average predictions across the 5 independently-trained models
  results <- bind_rows(test_preds_by_fold) %>%
    group_by(county_fips, county_name, state_po, vote_density, rep_share_early, rep_share_late, SHIFT) %>%
    summarise(Predicted_Shift = mean(Predicted_Shift), .groups = "drop") %>%
    mutate(Error = Predicted_Shift - SHIFT, Abs_Error = abs(Error))
  
  write_csv(results, out_name)
  cat("Saved ->", out_name, "\n")
}

# ==========================================
# 7. Execute All 9 Models
# ==========================================

cat("\n==========================================\n")
cat("EXECUTING SPATIAL MODELS (1-3)\n")
cat("==========================================\n")
# M1: Late Embeddings
run_spatial_change(data_1620, "16-20", "m1")
run_spatial_change(data_2024, "20-24", "m1")

# M2: Early Embeddings
run_spatial_change(data_1620, "16-20", "m2")
run_spatial_change(data_2024, "20-24", "m2")

# M3: Delta Embeddings
run_spatial_change(data_1620, "16-20", "m3")
run_spatial_change(data_2024, "20-24", "m3")

cat("\n==========================================\n")
cat("EXECUTING TEMPORAL MODELS (4-6)\n")
cat("==========================================\n")
# M4: Temporal (Late Embeddings)
run_temporal_change(data_1620, data_2024, "m4")

# M5: Temporal (Early Embeddings)
run_temporal_change(data_1620, data_2024, "m5")

# M6: Temporal (Delta Embeddings)
run_temporal_change(data_1620, data_2024, "m6")

cat("\nALL 9 MODELS SUCCESSFULLY TRAINED AND SAVED!\n")
```

What are the metrics for the change models?

```{r}
library(tidyverse)

# ==========================================
# 1. Define the 9 Output Files
# ==========================================

# Create a named vector mapping the exact file names to a readable description
model_files <- c(
  "Spatial M1 (16-20) | Late Emb"  = "data/results/change16-20_m1_results.csv",
  "Spatial M1 (20-24) | Late Emb"  = "data/results/change20-24_m1_results.csv",
  "Spatial M2 (16-20) | Early Emb" = "data/results/change16-20_m2_results.csv",
  "Spatial M2 (20-24) | Early Emb" = "data/results/change20-24_m2_results.csv",
  "Spatial M3 (16-20) | Delta Emb" = "data/results/change16-20_m3_results.csv",
  "Spatial M3 (20-24) | Delta Emb" = "data/results/change20-24_m3_results.csv",
  "Temporal M4 | Late Emb Transfer"  = "data/results/change_m4_results.csv",
  "Temporal M5 | Early Emb Transfer" = "data/results/change_m5_results.csv",
  "Temporal M6 | Delta Emb Transfer" = "data/results/change_m6_results.csv"
)

# ==========================================
# 2. Function to Calculate Metrics
# ==========================================

calculate_metrics <- function(file_path, model_name) {
  # Check if file exists to prevent crashing
  if (!file.exists(file_path)) {
    return(tibble(Model = model_name, R_Squared = NA, RMSE = NA, MAE = NA, Counties = 0))
  }
  
  # Load the results
  df <- read_csv(file_path, show_col_types = FALSE)
  
  # Calculate RSS, TSS, and final metrics
  rss  <- sum(df$Error^2)
  tss  <- sum((df$SHIFT - mean(df$SHIFT))^2)
  r2   <- 1 - (rss / tss)
  rmse <- sqrt(mean(df$Error^2))
  mae  <- mean(df$Abs_Error)
  
  # Return as a 1-row tibble
  tibble(
    Model = model_name,
    R_Squared = round(r2, 4),
    RMSE = round(rmse, 4),
    MAE = round(mae, 4),
    Counties = nrow(df)
  )
}

# ==========================================
# 3. Loop Through and Build Summary Table
# ==========================================

cat("Calculating metrics for all 9 models...\n")

# Use purrr to map over the named vector and bind the rows together
summary_table <- purrr::imap_dfr(model_files, ~ calculate_metrics(.x, .y))

# Print the final formatted table
cat("\n========================================================================\n")
cat("OVERALL MODEL PERFORMANCE SUMMARY\n")
cat("========================================================================\n")
print(as.data.frame(summary_table), row.names = FALSE)
cat("========================================================================\n")

```

## 5-Fold CV for 2016 (Predicting Trump Share)

```{r}

library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2016
clean_16 <- read_csv("data/prep/counties/clean_county_emb_16_17.csv") 

# Select features and remove rows with missing values
data_rf <- clean_16 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean") # Only embedding means
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

write_csv(full_results, "data/results/county16_m1_results.csv")

```

## 5-Fold CV for 2020 (Predicting Trump Share)

This script takes \~20 sec to run

```{r}

library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2020
clean_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv") 

# Select features and remove rows with missing values
data_rf <- clean_20 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean") # Only embedding means
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

write_csv(full_results, "data/results/county20_m1_results.csv")

```

R-Squared: 0.647 RMSE: 0.0958 Mean Abs Error: 0.0709

## 5-Fold CV for 2024 (Predicting Trump Share)

This script takes \~20 sec to run

```{r}

library(tidyverse)
library(ranger)
library(scales) # for comma labels in plots

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2024
clean_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv") 

# Select features and remove rows with missing values
data_rf <- clean_24 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean") # Only embedding means
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

# Save results 
write_csv(full_results, "data/results/county24_m1_results.csv")

```

R-Squared: 0.632 RMSE: 0.1189 Mean Abs Error: 0.094

## 5-FOLD CV, demographics 2020

```{r}

library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2020
clean_20 <- read_csv("data/prep/counties/master_county_emb_20.csv") 

# Select features and remove rows with missing values
data_rf <- clean_20 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean"), # Only embedding means
    pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, # Education
    unemployment_rate, civilian_labor_force # Labor
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

write_csv(full_results, "data/results/county20_m2_results.csv")

```

R-Squared: 0.742 RMSE: 0.0819 Mean Abs Error: 0.0596

## 5-Fold CV, demographics 2024

```{r}


library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2024
clean_24 <- read_csv("data/prep/counties/master_county_emb_24.csv") 

# Select features and remove rows with missing values
data_rf <- clean_24 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean"), # Only embedding means
    pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, # Education
    unemployment_rate, civilian_labor_force # Labor
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

write_csv(full_results, "data/results/county24_m2_results.csv")

```

R-Squared: 0.669 RMSE: 0.1127 Mean Abs Error: 0.0881

```{r}
summary(lm(rep_share ~ pct_less_than_hs + pct_hs_only + pct_some_college + pct_bachelors_plus + unemployment_rate + civilian_labor_force, data = clean_20))

summary(lm(rep_share ~ pct_less_than_hs + pct_hs_only + pct_some_college + pct_bachelors_plus + unemployment_rate + civilian_labor_force, data = clean_24))
```

### Visualizing results for county models

```{r}
# ==========================================
# 4. Overall Performance Metrics
# ==========================================

#full_results <- read.csv("data/results/county20_m1_results.csv")
full_results <- read.csv("data/results/county24_m2_results.csv")

# Calculate metrics on the full dataset (all folds combined)
rss  <- sum(full_results$Error^2)
tss  <- sum((full_results$TRUMPSHARE - mean(full_results$TRUMPSHARE))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(full_results$Error^2))

cat("\n==========================================")
cat("\nOVERALL 5-FOLD CV RESULTS")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4))
cat("\nMean Abs Error:", round(mean(full_results$Abs_Error), 4))
cat("\n==========================================\n")

# ==========================================
# 5. Plot 1: Prediction Error vs. Density
# ==========================================

plot1 <- ggplot(full_results, aes(x = vote_density, y = Error)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  scale_x_log10(labels = scales::comma) + 
  labs(
    title = "Prediction Error vs. Voter Density (5-Fold CV)",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 3)),
    x = "Voters per km² (Log Scale)",
    y = "Error (Predicted - Actual)"
  ) +
  theme_minimal()

print(plot1)

# ==========================================
# 6. Plot 2: R-squared per Density Decile
# ==========================================

decile_stats <- full_results %>%
  mutate(density_decile = ntile(vote_density, 10)) %>%
  group_by(density_decile) %>%
  summarize(
    avg_density = median(vote_density),
    n = n(),
    RSS = sum(Error^2),
    TSS = sum((TRUMPSHARE - mean(TRUMPSHARE))^2),
    # Safety check: avoid dividing by zero if TSS is 0
    R2 = if_else(TSS > 1e-6, 1 - (RSS / TSS), 0)
  )

plot2 <- ggplot(decile_stats, aes(x = factor(density_decile), y = R2)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(R2, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance by Density Decile",
    subtitle = "Decile 1 = Most Rural, Decile 10 = Most Urban",
    x = "Density Decile",
    y = "R-squared"
  ) +
  theme_minimal()

print(plot2)

# ==========================================
# 7. Plot 3: Predicted vs. Actual Trump Share
# ==========================================

plot3 <- ggplot(full_results, aes(x = TRUMPSHARE, y = Predicted_Share)) +
  # Add points with slight transparency
  geom_point(alpha = 0.4, size = 1, color = "darkslategrey") +
  
  # The 'Perfect Prediction' line (y = x)
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  
  # Add a trend line to see if the model is biased
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  
  # Format axes as percentages for readability
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  
  labs(
    title = "Actual vs. Predicted Trump Vote Share",
    subtitle = paste0("County-Level 5-Fold CV | R² = ", round(r2, 3)),
    x = "Actual Trump Share",
    y = "Predicted Trump Share",
    caption = "Red dashed line represents perfect prediction."
  ) +
  theme_minimal()

print(plot3)
```

```{r}

# Compare the "Spread" of the target variable
cat("2020 SD:", sd(clean_20$rep_share, na.rm=TRUE), "\n")
cat("2024 SD:", sd(clean_24$rep_share, na.rm=TRUE), "\n")

```

## 5-Fold CV for CHANGE in Trumpshare, 2024 change from 2020

This script takes \~20 sec to run

```{r}

# ==========================================
# 1. Load & Join Data (2020 vs 2024)
# ==========================================

# Load clean datasets
# Note: Added `total_votes_20 = total_votes` assuming your CSV has a total_votes column!
df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv") %>%
  select(county_fips, rep_share_20 = rep_share, total_votes_20 = total_votes)

df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv") %>%
  # We rename the 2024 share and total votes
  rename(rep_share_24 = rep_share, total_votes_24 = total_votes)

# Join the years to calculate the shift
data_shift <- inner_join(df_20, df_24, by = "county_fips") %>%
  mutate(
    # TARGET VARIABLE: Positive = Shift toward Trump, Negative = Shift toward Harris/Biden
    SHIFT_20_24 = rep_share_24 - rep_share_20,
    # NEW: Calculate the percentage change in total votes
    VOTE_CHANGE = (total_votes_24 - total_votes_20) / total_votes_20
  ) %>%
  # Prepare for modeling: Keep ID, Shift, 2020 Baseline, Vote Change, and Embeddings
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_20,    # Kept for analysis/plotting, NOT for training
    SHIFT_20_24,     # The Target
    VOTE_CHANGE,     # Kept temporarily for filtering
    ends_with("_mean") # The Predictors (2024 embeddings)
  ) %>%
  na.omit()

cat("Data Loaded. Analyzing shift for", nrow(data_shift), "counties.\n")
cat("Average Shift:", round(mean(data_shift$SHIFT_20_24) * 100, 2), "points\n")

# ==========================================
# FILTER: Remove extreme turnout changes using IQR (2x)
# ==========================================

# 1. Calculate IQR bounds for 2020-2024 Vote Change
q1 <- quantile(data_shift$VOTE_CHANGE, 0.25, na.rm = TRUE)
q3 <- quantile(data_shift$VOTE_CHANGE, 0.75, na.rm = TRUE)
iqr <- q3 - q1
lower_bound <- q1 - 2 * iqr
upper_bound <- q3 + 2 * iqr

# 2. Plot a scatterplot of change in voteshare vs change in total_votes
p_outliers <- ggplot(data_shift, aes(x = VOTE_CHANGE, y = SHIFT_20_24)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  # Add red lines to show exactly where we are cutting the data dynamically
  geom_vline(xintercept = c(lower_bound, upper_bound), linetype = "dashed", color = "red") +
  labs(
    title = "Vote Share Shift vs. Total Votes Change (2020-2024)",
    subtitle = sprintf("Red lines = 2*IQR bounds (%.1f%% to %.1f%%)", lower_bound*100, upper_bound*100),
    x = "Change in Total Votes (2020 to 2024)",
    y = "Shift in GOP Vote Share"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p_outliers)

cat("\n[Turnout IQR filter] Before:", nrow(data_shift), "counties\n")

# 3. Filter out counties outside the IQR bounds
data_filtered <- data_shift %>%
  filter(VOTE_CHANGE >= lower_bound & VOTE_CHANGE <= upper_bound)

# Print how many counties were removed
counties_removed <- nrow(data_shift) - nrow(data_filtered)

cat("[Turnout IQR filter] After: ", nrow(data_filtered), "counties\n")
cat("[Turnout IQR filter] Removed:", counties_removed, "counties due to extreme total vote swing\n")


# Create a dataframe of just the dropped counties
dropped_counties <- data_shift %>% # (assuming you have the pre-filter data)
  filter(VOTE_CHANGE < lower_bound | VOTE_CHANGE > upper_bound)

# Look at which states lost the most counties
table(dropped_counties$state_po) %>% sort(decreasing = TRUE) %>% head(10)

# 4. Reassign back to data_shift and drop VOTE_CHANGE so it doesn't leak into the Random Forest model
data_shift <- data_filtered %>% 
  select(-VOTE_CHANGE)

# Optional quick summary of remaining shift range
cat("[Turnout IQR filter] Vote Share Shift range now:",
    round(min(data_shift$SHIFT_20_24), 4), "to",
    round(max(data_shift$SHIFT_20_24), 4), "\n")


# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(55) # New seed for luck
data_shift <- data_shift %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

all_preds <- list()
importances <- list() # Store variable importance for each fold

cat("\nStarting CV to predict Swing/Shift...\n")

# ==========================================
# 3. CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split
  train_df <- data_shift %>% filter(fold_id != k)
  test_df  <- data_shift %>% filter(fold_id == k)
  
  # B. Train Model
  # Predict SHIFT using only Embeddings
  # Note: We exclude rep_share_20 to see if *Geography alone* predicts the swing
  rf_model <- ranger(
    SHIFT_20_24 ~ . - county_fips - county_name - state_po - vote_density - rep_share_20 - fold_id,
    data       = train_df,
    num.trees  = 500,
    mtry       = 21, # ~64 predictors / 3
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Predictions
  test_df$Predicted_Shift <- preds
  all_preds[[k]] <- test_df
  
  # E. Store Importance (optional, for curiosity)
  importances[[k]] <- rf_model$variable.importance
  
  cat("  Fold", k, "done.\n")
}

# Combine Results
results <- bind_rows(all_preds) %>%
  mutate(
    Error = Predicted_Shift - SHIFT_20_24,
    Abs_Error = abs(Error)
  )

# ==========================================
# 4. Metrics
# ==========================================

rss  <- sum(results$Error^2)
tss  <- sum((results$SHIFT_20_24 - mean(results$SHIFT_20_24))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(results$Error^2))

cat("\n==========================================")
cat("\nMODEL RESULTS: PREDICTING THE SWING")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4), "(", round(rmse*100, 2), "percentage points )")
cat("\n==========================================\n")

# ==========================================
# 5. Visualization 1: Actual vs. Predicted Shift
# ==========================================

# A 45-degree line plot to see if the model captures the direction of the shift
p1 <- ggplot(results, aes(x = SHIFT_20_24, y = Predicted_Shift)) +
  geom_point(alpha = 0.3, size = 1, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Can Satellites Predict Political Swings?",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 4)),
    x = "Actual Shift (2020 -> 2024)",
    y = "Predicted Shift (Based on Embeddings)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p1)

# ==========================================
# 6. Visualization 2: Error Bias by Baseline Partisanship
# ==========================================

# Did the model fail more in Deep Red or Deep Blue counties?
# We bin the 2020 Vote Share to see where the "surprises" happened.

p2 <- ggplot(results, aes(x = rep_share_20, y = Error)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", color = "blue", fill = "gray90") +
  labs(
    title = "Where were the surprises?",
    subtitle = "Positive Error = Model expected a more pro-Trump shift than happened",
    x = "2020 GOP Vote Share (Baseline)",
    y = "Prediction Error (Pred Swing - Actual Swing)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal()

print(p2)

# ==========================================
# 7. Visualization 3: Top Predictive Features
# ==========================================

# Average importance across folds
avg_imp <- bind_rows(lapply(importances, function(x) as.data.frame(t(x)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Importance") %>%
  arrange(desc(Importance)) %>%
  slice(1:15)

p3 <- ggplot(avg_imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Which Embeddings Predict the Swing?",
    subtitle = "Top 15 Features driving the change from 2020 to 2024",
    x = "",
    y = "Variable Importance (Impurity)"
  ) +
  theme_minimal()

print(p3)

# ==========================================
# 8. Visualization 4: R² by Vote Density Decile
# ==========================================

# Helper to compute R² safely (returns NA if variance is zero)
r2_fun <- function(actual, pred) {
  rss <- sum((pred - actual)^2, na.rm = TRUE)
  tss <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  if (is.na(tss) || tss == 0) return(NA_real_)
  1 - rss / tss
}

# Create density deciles + compute per-decile metrics
r2_by_decile <- results %>%
  mutate(
    density_decile = ntile(vote_density, 10)
  ) %>%
  group_by(density_decile) %>%
  summarise(
    n = n(),
    r2 = r2_fun(SHIFT_20_24, Predicted_Shift),
    rmse = sqrt(mean((Predicted_Shift - SHIFT_20_24)^2, na.rm = TRUE)),
    density_min = min(vote_density, na.rm = TRUE),
    density_max = max(vote_density, na.rm = TRUE),
    .groups = "drop"
  )

print(r2_by_decile)

# Bar plot of R² by decile
p4 <- ggplot(r2_by_decile, aes(x = factor(density_decile), y = r2)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Model Performance by Vote Density",
    subtitle = "Out-of-sample R² computed within each vote_density decile",
    x = "Vote Density Decile (1 = lowest density, 10 = highest)",
    y = "R² (within decile)"
  ) +
  theme_minimal()

print(p4)

write_csv(results, "data/results/county_change_m1_results.csv")


```

## Predicting only with the demographics

```{r}


library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2020
clean_20 <- read_csv("data/prep/counties/master_county_emb_20.csv") 

# Select features and remove rows with missing values
data_rf <- clean_20 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    #ends_with("_mean"), # Only embedding means
    pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, # Education
    unemployment_rate, civilian_labor_force # Labor
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

# ==========================================
# 4. Overall Performance Metrics
# ==========================================

#full_results <- read.csv("data/results/county20_m1_results.csv")
#full_results <- read.csv("data/results/county24_m2_results.csv")

# Calculate metrics on the full dataset (all folds combined)
rss  <- sum(full_results$Error^2)
tss  <- sum((full_results$TRUMPSHARE - mean(full_results$TRUMPSHARE))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(full_results$Error^2))

cat("\n==========================================")
cat("\nOVERALL 5-FOLD CV RESULTS")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4))
cat("\nMean Abs Error:", round(mean(full_results$Abs_Error), 4))
cat("\n==========================================\n")

# ==========================================
# 5. Plot 1: Prediction Error vs. Density
# ==========================================

plot1 <- ggplot(full_results, aes(x = vote_density, y = Error)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  scale_x_log10(labels = scales::comma) + 
  labs(
    title = "Prediction Error vs. Voter Density (5-Fold CV)",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 3)),
    x = "Voters per km² (Log Scale)",
    y = "Error (Predicted - Actual)"
  ) +
  theme_minimal()

print(plot1)

# ==========================================
# 6. Plot 2: R-squared per Density Decile
# ==========================================

decile_stats <- full_results %>%
  mutate(density_decile = ntile(vote_density, 10)) %>%
  group_by(density_decile) %>%
  summarize(
    avg_density = median(vote_density),
    n = n(),
    RSS = sum(Error^2),
    TSS = sum((TRUMPSHARE - mean(TRUMPSHARE))^2),
    # Safety check: avoid dividing by zero if TSS is 0
    R2 = if_else(TSS > 1e-6, 1 - (RSS / TSS), 0)
  )

plot2 <- ggplot(decile_stats, aes(x = factor(density_decile), y = R2)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(R2, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance by Density Decile",
    subtitle = "Decile 1 = Most Rural, Decile 10 = Most Urban",
    x = "Density Decile",
    y = "R-squared"
  ) +
  theme_minimal()

print(plot2)

# ==========================================
# 7. Plot 3: Predicted vs. Actual Trump Share
# ==========================================

plot3 <- ggplot(full_results, aes(x = TRUMPSHARE, y = Predicted_Share)) +
  # Add points with slight transparency
  geom_point(alpha = 0.4, size = 1, color = "darkslategrey") +
  
  # The 'Perfect Prediction' line (y = x)
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  
  # Add a trend line to see if the model is biased
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  
  # Format axes as percentages for readability
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  
  labs(
    title = "Actual vs. Predicted Trump Vote Share",
    subtitle = paste0("County-Level 5-Fold CV | R² = ", round(r2, 3)),
    x = "Actual Trump Share",
    y = "Predicted Trump Share",
    caption = "Red dashed line represents perfect prediction."
  ) +
  theme_minimal()

print(plot3)

```

train on 2016 -\> 2020, then predict 2020 -\> 2024 shifts, WITHOUT training on 2020 -\> 2024.

```{r}

library(tidyverse)
library(ranger)

# ==============================================================================
# GOAL:
#   Train on 2016 -> 2020 shifts (using 2020 embeddings as predictors),
#   then predict 2020 -> 2024 shifts (using 2024 embeddings as predictors),
#   WITHOUT training on 2020 -> 2024.
# ==============================================================================


# ==========================================
# 1. Load & Join Data for TRAIN (2016 vs 2020)
# ==========================================

df_16 <- read_csv("data/prep/counties/clean_county_emb_16_17.csv") %>%
  select(county_fips, rep_share_16 = rep_share, total_votes_16 = total_votes)

df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv") %>%
  rename(rep_share_20 = rep_share, total_votes_20 = total_votes)

# TRAIN set: 2016 -> 2020 shift, predictors = 2020 embeddings
train_shift <- inner_join(df_16, df_20, by = "county_fips") %>%
  mutate(
    SHIFT_16_20 = rep_share_20 - rep_share_16,
    VOTE_CHANGE_16_20 = (total_votes_20 - total_votes_16) / total_votes_16
  ) %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_16,
    rep_share_20,      # kept for analysis, not used for training
    SHIFT_16_20,       # target for training
    VOTE_CHANGE_16_20, # kept temporarily for filtering
    ends_with("_mean") # predictors = 2020 embeddings
  ) %>%
  na.omit()

cat("TRAIN data loaded (2016->2020) for", nrow(train_shift), "counties.\n")
cat("Avg TRAIN shift:", round(mean(train_shift$SHIFT_16_20) * 100, 2), "points\n")


# ==========================================
# 2. Load & Join Data for TEST (2020 vs 2024)
# ==========================================

df_20_base <- read_csv("data/prep/counties/clean_county_emb_20.csv") %>%
  select(county_fips, rep_share_20 = rep_share, total_votes_20 = total_votes)

df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv") %>%
  rename(rep_share_24 = rep_share, total_votes_24 = total_votes)

# TEST set: 2020 -> 2024 shift, predictors = 2024 embeddings
test_shift <- inner_join(df_20_base, df_24, by = "county_fips") %>%
  mutate(
    SHIFT_20_24 = rep_share_24 - rep_share_20,
    VOTE_CHANGE_20_24 = (total_votes_24 - total_votes_20) / total_votes_20
  ) %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_20,      # kept for analysis, not used for training
    SHIFT_20_24,       # target only for evaluation
    VOTE_CHANGE_20_24, # kept temporarily for filtering
    ends_with("_mean") # predictors = 2024 embeddings
  ) %>%
  na.omit()

cat("\nTEST data loaded (2020->2024) for", nrow(test_shift), "counties.\n")
cat("Avg TEST shift:", round(mean(test_shift$SHIFT_20_24) * 100, 2), "points\n")

# ==========================================
# FILTER: Remove extreme turnout changes using IQR (2x)
# ==========================================

# --- TRAIN outliers (2016 -> 2020) ---
# 1. Calculate IQR bounds for TRAIN
q1_train <- quantile(train_shift$VOTE_CHANGE_16_20, 0.25, na.rm = TRUE)
q3_train <- quantile(train_shift$VOTE_CHANGE_16_20, 0.75, na.rm = TRUE)
iqr_train <- q3_train - q1_train
lower_train <- q1_train - 2 * iqr_train
upper_train <- q3_train + 2 * iqr_train

# 2. Plot TRAIN with dynamic bounds
p_outliers_train <- ggplot(train_shift, aes(x = VOTE_CHANGE_16_20, y = SHIFT_16_20)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_vline(xintercept = c(lower_train, upper_train), linetype = "dashed", color = "red") +
  labs(
    title = "TRAIN: Vote Share Shift vs. Total Votes Change (2016-2020)",
    subtitle = sprintf("Red lines = 2*IQR bounds (%.1f%% to %.1f%%)", lower_train*100, upper_train*100),
    x = "Change in Total Votes (2016 to 2020)",
    y = "Shift in GOP Vote Share"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p_outliers_train)

# 3. Filter TRAIN
n_train_before <- nrow(train_shift)
train_shift <- train_shift %>% 
  filter(VOTE_CHANGE_16_20 >= lower_train & VOTE_CHANGE_16_20 <= upper_train)
n_train_after <- nrow(train_shift)

cat("\n[Turnout IQR filter | TRAIN] Before:", n_train_before,
    "| After:", n_train_after,
    "| Removed:", (n_train_before - n_train_after), "counties\n")

# Drop temporary column
train_shift <- train_shift %>% select(-VOTE_CHANGE_16_20)


# --- TEST outliers (2020 -> 2024) ---
# 1. Calculate IQR bounds for TEST
q1_test <- quantile(test_shift$VOTE_CHANGE_20_24, 0.25, na.rm = TRUE)
q3_test <- quantile(test_shift$VOTE_CHANGE_20_24, 0.75, na.rm = TRUE)
iqr_test <- q3_test - q1_test
lower_test <- q1_test - 2 * iqr_test
upper_test <- q3_test + 2 * iqr_test

# 2. Plot TEST with dynamic bounds
p_outliers_test <- ggplot(test_shift, aes(x = VOTE_CHANGE_20_24, y = SHIFT_20_24)) +
  geom_point(alpha = 0.5, color = "darkorange") +
  geom_vline(xintercept = c(lower_test, upper_test), linetype = "dashed", color = "red") +
  labs(
    title = "TEST: Vote Share Shift vs. Total Votes Change (2020-2024)",
    subtitle = sprintf("Red lines = 2*IQR bounds (%.1f%% to %.1f%%)", lower_test*100, upper_test*100),
    x = "Change in Total Votes (2020 to 2024)",
    y = "Shift in GOP Vote Share"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p_outliers_test)

# 3. Filter TEST
n_test_before <- nrow(test_shift)
test_shift <- test_shift %>% 
  filter(VOTE_CHANGE_20_24 >= lower_test & VOTE_CHANGE_20_24 <= upper_test)
n_test_after <- nrow(test_shift)

cat("\n[Turnout IQR filter | TEST]  Before:", n_test_before,
    "| After:", n_test_after,
    "| Removed:", (n_test_before - n_test_after), "counties\n")

# Drop temporary column
test_shift <- test_shift %>% select(-VOTE_CHANGE_20_24)

# ==========================================
# 3. Make sure TRAIN and TEST have identical predictor columns
# ==========================================

id_cols_train <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_16", "rep_share_20", "SHIFT_16_20")
id_cols_test  <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_20", "SHIFT_20_24")

x_cols_train <- setdiff(names(train_shift), id_cols_train)
x_cols_test  <- setdiff(names(test_shift),  id_cols_test)

common_x <- intersect(x_cols_train, x_cols_test)

# Keep only the shared predictors (safe if columns differ across years)
train_shift <- train_shift %>% select(all_of(id_cols_train), all_of(common_x))
test_shift  <- test_shift  %>% select(all_of(id_cols_test),  all_of(common_x))

cat("\nUsing", length(common_x), "shared embedding predictors.\n")


# ==========================================
# 4. 5-Fold CV ON TRAIN (2016->2020), predict TEST (2020->2024)
# ==========================================

set.seed(55)
train_shift <- train_shift %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

test_preds_by_fold <- list()
importances <- list()

cat("\nStarting CV: train on 2016->2020, predict 2020->2024...\n")

for (k in 1:5) {

  # A. Train split
  train_df <- train_shift %>% filter(fold_id != k)

  # B. Fit model on 2016->2020 shift
  rf_model <- ranger(
    SHIFT_16_20 ~ . - county_fips - county_name - state_po - vote_density - rep_share_16 - rep_share_20 - fold_id,
    data       = train_df,
    num.trees  = 500,
    mtry       = 21,           # keep your setting; tweak if you want
    importance = "impurity",
    seed       = 42
  )

  # C. Predict 2020->2024 shifts for ALL test counties using this fold's model
  preds <- predict(rf_model, data = test_shift)$predictions

  tmp <- test_shift %>%
    mutate(
      fold_model = k,
      Predicted_Shift = preds
    )

  test_preds_by_fold[[k]] <- tmp
  importances[[k]] <- rf_model$variable.importance

  cat("  Fold", k, "model trained (on 2016->2020) and applied to 2020->2024.\n")
}

# Average predictions across the 5 independently-trained models
results <- bind_rows(test_preds_by_fold) %>%
  group_by(county_fips, county_name, state_po, vote_density, rep_share_20, SHIFT_20_24) %>%
  summarise(
    Predicted_Shift = mean(Predicted_Shift),
    .groups = "drop"
  ) %>%
  mutate(
    Error = Predicted_Shift - SHIFT_20_24,
    Abs_Error = abs(Error)
  )


# ==========================================
# 5. Metrics (evaluated on 2020->2024)
# ==========================================

rss  <- sum(results$Error^2)
tss  <- sum((results$SHIFT_20_24 - mean(results$SHIFT_20_24))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(results$Error^2))

cat("\n==========================================")
cat("\nTRANSFER MODEL RESULTS")
cat("\nTrain: 2016->2020 | Test: 2020->2024")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4), "(", round(rmse*100, 2), "percentage points )")
cat("\n==========================================\n")


# ==========================================
# 6. Visualization 1: Actual vs. Predicted Shift
# ==========================================

p1 <- ggplot(results, aes(x = SHIFT_20_24, y = Predicted_Shift)) +
  geom_point(alpha = 0.3, size = 1, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Out-of-period generalization: 2016→2020 trained, 2020→2024 predicted",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 4)),
    x = "Actual Shift (2020 → 2024)",
    y = "Predicted Shift (trained on 2016 → 2020)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p1)


# ==========================================
# 7. Visualization 2: Error Bias by Baseline Partisanship
# ==========================================

p2 <- ggplot(results, aes(x = rep_share_20, y = Error)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", color = "blue", fill = "gray90") +
  labs(
    title = "Where were the surprises (transfer model)?",
    subtitle = "Positive Error = model expected a more pro-GOP shift than happened",
    x = "2020 GOP Vote Share (Baseline)",
    y = "Prediction Error (Pred - Actual)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal()

print(p2)


# ==========================================
# 8. Visualization 3: Top Predictive Features (avg across folds)
# ==========================================

avg_imp <- bind_rows(lapply(importances, function(x) as.data.frame(t(x)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Importance") %>%
  arrange(desc(Importance)) %>%
  slice(1:15)

p3 <- ggplot(avg_imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Which embeddings predict swings (learned from 2016→2020)?",
    subtitle = "Top 15 features by impurity importance (avg across folds)",
    x = "",
    y = "Variable Importance (Impurity)"
  ) +
  theme_minimal()

print(p3)


# ==========================================
# 9. Visualization 4: R² by Vote Density Decile (TRANSFER MODEL)
#   Drop this right after p3 (and before write_csv).
# ==========================================

# Helper to compute R² safely (returns NA if variance is zero)
r2_fun <- function(actual, pred) {
  rss <- sum((pred - actual)^2, na.rm = TRUE)
  tss <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  if (is.na(tss) || tss == 0) return(NA_real_)
  1 - rss / tss
}

r2_by_density_decile <- results %>%
  mutate(
    density_decile = ntile(vote_density, 10)
    # If density is super skewed, you may prefer:
    # density_decile = ntile(log1p(vote_density), 10)
  ) %>%
  group_by(density_decile) %>%
  summarise(
    n = n(),
    r2 = r2_fun(SHIFT_20_24, Predicted_Shift),
    rmse = sqrt(mean((Predicted_Shift - SHIFT_20_24)^2, na.rm = TRUE)),
    density_min = min(vote_density, na.rm = TRUE),
    density_max = max(vote_density, na.rm = TRUE),
    .groups = "drop"
  )

print(r2_by_density_decile)

p4 <- ggplot(r2_by_density_decile, aes(x = factor(density_decile), y = r2)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Transfer-model performance by vote density",
    subtitle = "R² computed within each vote_density decile (test period: 2020→2024)",
    x = "Vote Density Decile (1 = lowest density, 10 = highest)",
    y = "R² (within decile)"
  ) +
  theme_minimal()

print(p4)

# ==========================================
# 9. Save
# ==========================================

write_csv(results, "data/results/county_change_m2_results.csv")
cat("Saved")



```

```{r}

## CHANGE to voter turnout outliers

library(tidyverse)
library(ranger)

# ==============================================================================
# GOAL (EARLY-YEAR EMBEDDINGS):
#   Train on 2016 -> 2020 shifts using **2016 embeddings** as predictors,
#   then predict 2020 -> 2024 shifts using **2020 embeddings** as predictors,
#   WITHOUT training on 2020 -> 2024.
# ==============================================================================


# ==========================================
# 1. Load Data
# ==========================================

# Full files (keep embeddings/metadata)
df_16_full <- read_csv("data/prep/counties/clean_county_emb_16_17.csv")
df_20_full <- read_csv("data/prep/counties/clean_county_emb_20.csv")
df_24_full <- read_csv("data/prep/counties/clean_county_emb_24.csv")

# Shares-only extracts (to avoid duplicate cols in joins)
df_16_share <- df_16_full %>% select(county_fips, rep_share_16 = rep_share)
df_20_share <- df_20_full %>% select(county_fips, rep_share_20 = rep_share)
df_24_share <- df_24_full %>% select(county_fips, rep_share_24 = rep_share)


# ==========================================
# 2. TRAIN: 2016 -> 2020 shift, predictors = 2016 embeddings
# ==========================================

train_shift <- inner_join(df_16_share, df_20_share, by = "county_fips") %>%
  # attach 2016 embeddings + metadata we want to keep
  inner_join(
    df_16_full %>%
      select(county_fips, county_name, state_po, vote_density, ends_with("_mean")),
    by = "county_fips"
  ) %>%
  mutate(
    SHIFT_16_20 = rep_share_20 - rep_share_16
  ) %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_16,
    rep_share_20,      # kept for analysis, not used for training
    SHIFT_16_20,       # target for training
    ends_with("_mean") # predictors = 2016 embeddings
  ) %>%
  na.omit()

cat("TRAIN data loaded (2016->2020) for", nrow(train_shift), "counties.\n")
cat("Avg TRAIN shift:", round(mean(train_shift$SHIFT_16_20) * 100, 2), "points\n")


# ==========================================
# 3. TEST: 2020 -> 2024 shift, predictors = 2020 embeddings
# ==========================================

test_shift <- inner_join(df_20_share, df_24_share, by = "county_fips") %>%
  # attach 2020 embeddings + metadata we want to keep
  inner_join(
    df_20_full %>%
      select(county_fips, county_name, state_po, vote_density, ends_with("_mean")),
    by = "county_fips"
  ) %>%
  mutate(
    SHIFT_20_24 = rep_share_24 - rep_share_20
  ) %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_20,      # kept for analysis, not used for training
    SHIFT_20_24,       # target only for evaluation
    ends_with("_mean") # predictors = 2020 embeddings
  ) %>%
  na.omit()

cat("\nTEST data loaded (2020->2024) for", nrow(test_shift), "counties.\n")
cat("Avg TEST shift:", round(mean(test_shift$SHIFT_20_24) * 100, 2), "points\n")


# ==========================================
# OPTIONAL: Remove extreme shift outliers (|SHIFT| > 0.15)
#   Drop this RIGHT AFTER you create train_shift and test_shift (after na.omit()).
#   Easy to comment out as a block.
# ==========================================

outlier_cutoff <- 0.15

# --- TRAIN outliers (2016 -> 2020) ---
n_train_before <- nrow(train_shift)
train_shift <- train_shift %>% filter(abs(SHIFT_16_20) <= outlier_cutoff)
n_train_after <- nrow(train_shift)

cat("\n[Outlier filter | TRAIN] Before:", n_train_before,
    "| After:", n_train_after,
    "| Removed:", (n_train_before - n_train_after), "\n")
cat("[Outlier filter | TRAIN] Shift range now:",
    round(min(train_shift$SHIFT_16_20), 4), "to",
    round(max(train_shift$SHIFT_16_20), 4), "\n")

# --- TEST outliers (2020 -> 2024) ---
n_test_before <- nrow(test_shift)
test_shift <- test_shift %>% filter(abs(SHIFT_20_24) <= outlier_cutoff)
n_test_after <- nrow(test_shift)

cat("\n[Outlier filter | TEST]  Before:", n_test_before,
    "| After:", n_test_after,
    "| Removed:", (n_test_before - n_test_after), "\n")
cat("[Outlier filter | TEST]  Shift range now:",
    round(min(test_shift$SHIFT_20_24), 4), "to",
    round(max(test_shift$SHIFT_20_24), 4), "\n")


# ==========================================
# 4. Make sure TRAIN and TEST have identical predictor columns
# ==========================================

id_cols_train <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_16", "rep_share_20", "SHIFT_16_20")
id_cols_test  <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_20", "SHIFT_20_24")

x_cols_train <- setdiff(names(train_shift), id_cols_train)
x_cols_test  <- setdiff(names(test_shift),  id_cols_test)

common_x <- intersect(x_cols_train, x_cols_test)

train_shift <- train_shift %>% select(all_of(id_cols_train), all_of(common_x))
test_shift  <- test_shift  %>% select(all_of(id_cols_test),  all_of(common_x))

cat("\nUsing", length(common_x), "shared embedding predictors.\n")


# ==========================================
# 5. 5-Fold CV ON TRAIN (2016->2020), predict TEST (2020->2024)
# ==========================================

set.seed(55)
train_shift <- train_shift %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

test_preds_by_fold <- list()
importances <- list()

cat("\nStarting CV: train on 2016->2020 (2016 embeddings), predict 2020->2024 (2020 embeddings)...\n")

for (k in 1:5) {

  train_df <- train_shift %>% filter(fold_id != k)

  rf_model <- ranger(
    SHIFT_16_20 ~ . - county_fips - county_name - state_po - vote_density - rep_share_16 - rep_share_20 - fold_id,
    data       = train_df,
    num.trees  = 500,
    mtry       = 21,
    importance = "impurity",
    seed       = 42
  )

  preds <- predict(rf_model, data = test_shift)$predictions

  tmp <- test_shift %>%
    mutate(
      fold_model = k,
      Predicted_Shift = preds
    )

  test_preds_by_fold[[k]] <- tmp
  importances[[k]] <- rf_model$variable.importance

  cat("  Fold", k, "model trained (2016->2020) and applied to 2020->2024.\n")
}

results <- bind_rows(test_preds_by_fold) %>%
  group_by(county_fips, county_name, state_po, vote_density, rep_share_20, SHIFT_20_24) %>%
  summarise(
    Predicted_Shift = mean(Predicted_Shift),
    .groups = "drop"
  ) %>%
  mutate(
    Error = Predicted_Shift - SHIFT_20_24,
    Abs_Error = abs(Error)
  )


# ==========================================
# 6. Metrics (evaluated on 2020->2024)
# ==========================================

rss  <- sum(results$Error^2)
tss  <- sum((results$SHIFT_20_24 - mean(results$SHIFT_20_24))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(results$Error^2))

cat("\n==========================================")
cat("\nTRANSFER MODEL RESULTS (EARLY-YEAR EMBEDDINGS)")
cat("\nTrain: 2016->2020 (2016 emb) | Test: 2020->2024 (2020 emb)")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4), "(", round(rmse*100, 2), "percentage points )")
cat("\n==========================================\n")


# ==========================================
# 7. Visualizations (same as before)
# ==========================================

p1 <- ggplot(results, aes(x = SHIFT_20_24, y = Predicted_Shift)) +
  geom_point(alpha = 0.3, size = 1, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Early-year embeddings: 2016→2020 trained, 2020→2024 predicted",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 4)),
    x = "Actual Shift (2020 → 2024)",
    y = "Predicted Shift (trained on 2016 → 2020)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
print(p1)

p2 <- ggplot(results, aes(x = rep_share_20, y = Error)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", color = "blue", fill = "gray90") +
  labs(
    title = "Where were the surprises (transfer model, early embeddings)?",
    subtitle = "Positive Error = model expected a more pro-GOP shift than happened",
    x = "2020 GOP Vote Share (Baseline)",
    y = "Prediction Error (Pred - Actual)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal()
print(p2)

avg_imp <- bind_rows(lapply(importances, function(x) as.data.frame(t(x)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Importance") %>%
  arrange(desc(Importance)) %>%
  slice(1:15)

p3 <- ggplot(avg_imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Top features (learned from 2016 embeddings)",
    subtitle = "Top 15 features by impurity importance (avg across folds)",
    x = "",
    y = "Variable Importance (Impurity)"
  ) +
  theme_minimal()
print(p3)

# ==========================================
# 9. Visualization 4: R² by Vote Density Decile (TRANSFER MODEL)
#   Drop this right after p3 (and before write_csv).
# ==========================================

# Helper to compute R² safely (returns NA if variance is zero)
r2_fun <- function(actual, pred) {
  rss <- sum((pred - actual)^2, na.rm = TRUE)
  tss <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  if (is.na(tss) || tss == 0) return(NA_real_)
  1 - rss / tss
}

r2_by_density_decile <- results %>%
  mutate(
    density_decile = ntile(vote_density, 10)
    # If density is super skewed, you may prefer:
    # density_decile = ntile(log1p(vote_density), 10)
  ) %>%
  group_by(density_decile) %>%
  summarise(
    n = n(),
    r2 = r2_fun(SHIFT_20_24, Predicted_Shift),
    rmse = sqrt(mean((Predicted_Shift - SHIFT_20_24)^2, na.rm = TRUE)),
    density_min = min(vote_density, na.rm = TRUE),
    density_max = max(vote_density, na.rm = TRUE),
    .groups = "drop"
  )

print(r2_by_density_decile)

p4 <- ggplot(r2_by_density_decile, aes(x = factor(density_decile), y = r2)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Transfer-model performance by vote density",
    subtitle = "R² computed within each vote_density decile (test period: 2020→2024)",
    x = "Vote Density Decile (1 = lowest density, 10 = highest)",
    y = "R² (within decile)"
  ) +
  theme_minimal()

print(p4)

# ==========================================
# 8. Save
# ==========================================

write_csv(results, "data/results/county_change_m2_early_embeddings_results.csv")
cat("Saved: data/results/county_change_m2_early_embeddings_results.csv\n")
```

```{r}

## CHANGE to voter turnout outliers

library(tidyverse)
library(ranger)

# ==============================================================================
# GOAL (DELTA MODEL):
#   Train on 2016 -> 2020 shifts using DELTA embeddings (2020 - 2016),
#   then predict 2020 -> 2024 shifts using DELTA embeddings (2024 - 2020),
#   WITHOUT training on 2020 -> 2024 shifts.
#
#   Predictors: Axx_mean deltas only (suffix: _delta)
#   Target (train): SHIFT_16_20
#   Eval (test):    SHIFT_20_24
# ==============================================================================

# ==========================================
# 0. Load base files
# ==========================================

df_16 <- read_csv("data/prep/counties/clean_county_emb_16_17.csv", show_col_types = FALSE)
df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv",     show_col_types = FALSE)
df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv",     show_col_types = FALSE)

# We'll use only the _mean embeddings for deltas (consistent with your earlier scripts)
emb_cols_16 <- names(df_16) %>% keep(~ str_detect(.x, "_mean$"))
emb_cols_20 <- names(df_20) %>% keep(~ str_detect(.x, "_mean$"))
emb_cols_24 <- names(df_24) %>% keep(~ str_detect(.x, "_mean$"))

common_emb <- Reduce(intersect, list(emb_cols_16, emb_cols_20, emb_cols_24))

cat("Using", length(common_emb), "shared *_mean embedding columns for deltas.\n")

# ==========================================
# 1. TRAIN data: 2016 vs 2020 (delta = 2020 - 2016)
# ==========================================

df_16_small <- df_16 %>%
  select(
    county_fips,
    rep_share_16 = rep_share,
    all_of(common_emb)
  ) %>%
  rename_with(~ paste0(.x, "_16"), all_of(common_emb))

df_20_small <- df_20 %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_20 = rep_share,
    all_of(common_emb)
  ) %>%
  rename_with(~ paste0(.x, "_20"), all_of(common_emb))

train_join <- inner_join(df_16_small, df_20_small, by = "county_fips") %>%
  mutate(SHIFT_16_20 = rep_share_20 - rep_share_16)

# Build delta predictors safely (NO cur_column)
train_deltas <- purrr::map_dfc(common_emb, function(nm) {
  train_join[[paste0(nm, "_20")]] - train_join[[paste0(nm, "_16")]]
})
names(train_deltas) <- paste0(common_emb, "_delta")

train_shift <- bind_cols(train_join, train_deltas) %>%
  select(
    county_fips, county_name, state_po, vote_density,
    rep_share_16, rep_share_20,
    SHIFT_16_20,
    ends_with("_delta")
  ) %>%
  na.omit()

cat("\nTRAIN data loaded (2016->2020) for", nrow(train_shift), "counties.\n")
cat("Avg TRAIN shift:", round(mean(train_shift$SHIFT_16_20) * 100, 2), "points\n")

# ==========================================
# 2. TEST data: 2020 vs 2024 (delta = 2024 - 2020)
# ==========================================

df_20_base <- df_20 %>%
  select(
    county_fips,
    rep_share_20 = rep_share,
    all_of(common_emb)
  ) %>%
  rename_with(~ paste0(.x, "_20"), all_of(common_emb))

df_24_small <- df_24 %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_24 = rep_share,
    all_of(common_emb)
  ) %>%
  rename_with(~ paste0(.x, "_24"), all_of(common_emb))

test_join <- inner_join(df_20_base, df_24_small, by = "county_fips") %>%
  mutate(SHIFT_20_24 = rep_share_24 - rep_share_20)

# Build delta predictors safely (NO cur_column)
test_deltas <- purrr::map_dfc(common_emb, function(nm) {
  test_join[[paste0(nm, "_24")]] - test_join[[paste0(nm, "_20")]]
})
names(test_deltas) <- paste0(common_emb, "_delta")

test_shift <- bind_cols(test_join, test_deltas) %>%
  select(
    county_fips, county_name, state_po, vote_density,
    rep_share_20,
    SHIFT_20_24,
    ends_with("_delta")
  ) %>%
  na.omit()

cat("\nTEST data loaded (2020->2024) for", nrow(test_shift), "counties.\n")
cat("Avg TEST shift:", round(mean(test_shift$SHIFT_20_24) * 100, 2), "points\n")

# ==========================================
# OPTIONAL: Remove extreme shift outliers (|SHIFT| > 0.15)
#   Drop this RIGHT AFTER you create train_shift and test_shift (after na.omit()).
#   Easy to comment out as a block.
# ==========================================

outlier_cutoff <- 0.15

# --- TRAIN outliers (2016 -> 2020) ---
n_train_before <- nrow(train_shift)
train_shift <- train_shift %>% filter(abs(SHIFT_16_20) <= outlier_cutoff)
n_train_after <- nrow(train_shift)

cat("\n[Outlier filter | TRAIN] Before:", n_train_before,
    "| After:", n_train_after,
    "| Removed:", (n_train_before - n_train_after), "\n")
cat("[Outlier filter | TRAIN] Shift range now:",
    round(min(train_shift$SHIFT_16_20), 4), "to",
    round(max(train_shift$SHIFT_16_20), 4), "\n")

# --- TEST outliers (2020 -> 2024) ---
n_test_before <- nrow(test_shift)
test_shift <- test_shift %>% filter(abs(SHIFT_20_24) <= outlier_cutoff)
n_test_after <- nrow(test_shift)

cat("\n[Outlier filter | TEST]  Before:", n_test_before,
    "| After:", n_test_after,
    "| Removed:", (n_test_before - n_test_after), "\n")
cat("[Outlier filter | TEST]  Shift range now:",
    round(min(test_shift$SHIFT_20_24), 4), "to",
    round(max(test_shift$SHIFT_20_24), 4), "\n")

# ==========================================
# 3. Ensure identical predictor columns between TRAIN and TEST
# ==========================================

id_cols_train <- c("county_fips", "county_name", "state_po", "vote_density",
                   "rep_share_16", "rep_share_20", "SHIFT_16_20")
id_cols_test  <- c("county_fips", "county_name", "state_po", "vote_density",
                   "rep_share_20", "SHIFT_20_24")

x_cols_train <- setdiff(names(train_shift), id_cols_train)
x_cols_test  <- setdiff(names(test_shift),  id_cols_test)

common_x <- intersect(x_cols_train, x_cols_test)

train_shift <- train_shift %>% select(all_of(id_cols_train), all_of(common_x))
test_shift  <- test_shift  %>% select(all_of(id_cols_test),  all_of(common_x))

cat("\nUsing", length(common_x), "shared DELTA predictors.\n")

# ==========================================
# 4. 5-Fold CV on TRAIN (2016->2020), predict TEST (2020->2024)
# ==========================================

set.seed(55)
train_shift <- train_shift %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

test_preds_by_fold <- list()
importances <- list()

cat("\nStarting CV: train on 2016->2020 DELTAS, predict 2020->2024 DELTAS...\n")

for (k in 1:5) {

  train_df <- train_shift %>% filter(fold_id != k)

  rf_model <- ranger(
    SHIFT_16_20 ~ . - county_fips - county_name - state_po - vote_density - rep_share_16 - rep_share_20 - fold_id,
    data       = train_df,
    num.trees  = 500,
    mtry       = min(21, length(common_x)),  # guard if fewer predictors
    importance = "impurity",
    seed       = 42
  )

  preds <- predict(rf_model, data = test_shift)$predictions

  tmp <- test_shift %>%
    mutate(
      fold_model = k,
      Predicted_Shift = preds
    )

  test_preds_by_fold[[k]] <- tmp
  importances[[k]] <- rf_model$variable.importance

  cat("  Fold", k, "model trained (on 2016->2020) and applied to 2020->2024.\n")
}

# Average predictions across the 5 independently-trained models
results <- bind_rows(test_preds_by_fold) %>%
  group_by(county_fips, county_name, state_po, vote_density, rep_share_20, SHIFT_20_24) %>%
  summarise(
    Predicted_Shift = mean(Predicted_Shift),
    .groups = "drop"
  ) %>%
  mutate(
    Error = Predicted_Shift - SHIFT_20_24,
    Abs_Error = abs(Error)
  )

# ==========================================
# 5. Metrics (evaluated on 2020->2024)
# ==========================================

rss  <- sum(results$Error^2)
tss  <- sum((results$SHIFT_20_24 - mean(results$SHIFT_20_24))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(results$Error^2))

cat("\n==========================================")
cat("\nDELTA TRANSFER MODEL RESULTS")
cat("\nTrain: 2016->2020 (delta 20-16) | Test: 2020->2024 (delta 24-20)")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4), "(", round(rmse*100, 2), "percentage points )")
cat("\n==========================================\n")

# ==========================================
# 6. Visualization 1: Actual vs Predicted Shift
# ==========================================

p1 <- ggplot(results, aes(x = SHIFT_20_24, y = Predicted_Shift)) +
  geom_point(alpha = 0.3, size = 1, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Delta-embeddings transfer: (20-16) trained, (24-20) predicted",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 4)),
    x = "Actual Shift (2020 → 2024)",
    y = "Predicted Shift (trained on 2016 → 2020 deltas)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p1)

# ==========================================
# 7. Visualization 2: Error vs Baseline Partisanship
# ==========================================

p2 <- ggplot(results, aes(x = rep_share_20, y = Error)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", color = "blue", fill = "gray90") +
  labs(
    title = "Where were the surprises (delta transfer model)?",
    subtitle = "Positive Error = model expected a more pro-GOP shift than happened",
    x = "2020 GOP Vote Share (Baseline)",
    y = "Prediction Error (Pred - Actual)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal()

print(p2)

# ==========================================
# 8. Visualization 3: Top Predictive Delta Features
# ==========================================

avg_imp <- bind_rows(lapply(importances, function(x) as.data.frame(t(x)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Importance") %>%
  arrange(desc(Importance)) %>%
  slice(1:15)

p3 <- ggplot(avg_imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Which delta embeddings predict swings (learned from 2016→2020)?",
    subtitle = "Top 15 delta features by impurity importance (avg across folds)",
    x = "",
    y = "Variable Importance (Impurity)"
  ) +
  theme_minimal()

print(p3)

# ==========================================
# 9. OPTIONAL DROP-IN VIS: R² by vote_density decile
#   - Put this AFTER you create `results` and compute r2/rmse (anywhere after results exists).
# ==========================================

results_r2_by_decile <- results %>%
  mutate(
    density_decile = ntile(vote_density, 10)
  ) %>%
  group_by(density_decile) %>%
  summarise(
    n = n(),
    rss = sum(Error^2),
    tss = sum((SHIFT_20_24 - mean(SHIFT_20_24))^2),
    r2  = 1 - rss / tss,
    rmse = sqrt(mean(Error^2)),
    density_min = min(vote_density, na.rm = TRUE),
    density_max = max(vote_density, na.rm = TRUE),
    .groups = "drop"
  )

p4 <- ggplot(results_r2_by_decile, aes(x = factor(density_decile), y = r2)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Model performance by vote-density decile (Delta transfer model)",
    subtitle = "R² computed within each decile on the 2020→2024 test set",
    x = "Vote-density decile (1 = lowest density, 10 = highest)",
    y = "R² within decile"
  ) +
  theme_minimal()

print(p4)

# ==========================================
# 10. Save
# ==========================================

write_csv(results, "data/results/county_change_m3_delta_results.csv")
cat("Saved: data/results/county_change_m3_delta_results.csv\n")

```

# Let's validate the data!

```{r}

library(tidyverse)

df_16 <- read.csv("data/prep/counties/clean_county_emb_16_17.csv")
df_20 <- read.csv("data/prep/counties/clean_county_emb_20.csv")
df_24 <- read.csv("data/prep/counties/clean_county_emb_24.csv")

dups <- function(df, name) {
  n_dup <- df %>% count(county_fips) %>% filter(n > 1) %>% nrow()
  cat(name, "- duplicate county_fips rows:", n_dup, "\n")
}

dups(df_16, "df_16")
dups(df_20, "df_20")
dups(df_24, "df_24")

train_join <- inner_join(
  df_16 %>% select(county_fips, rep_share_16 = rep_share),
  df_20 %>% select(county_fips, rep_share_20 = rep_share),
  by = "county_fips"
)

test_join <- inner_join(
  df_20 %>% select(county_fips, rep_share_20 = rep_share),
  df_24 %>% select(county_fips, rep_share_24 = rep_share),
  by = "county_fips"
)

cat("Rows train_join:", nrow(train_join), "\n")
cat("Unique county_fips train_join:", n_distinct(train_join$county_fips), "\n\n")

cat("Rows test_join:", nrow(test_join), "\n")
cat("Unique county_fips test_join:", n_distinct(test_join$county_fips), "\n")

```

```{r}

summary(df_16$rep_share)
summary(df_20$rep_share)
summary(df_24$rep_share)

range(df_16$rep_share, na.rm = TRUE)
range(df_20$rep_share, na.rm = TRUE)
range(df_24$rep_share, na.rm = TRUE)

```

```{r}
train_shift_vals <- train_join %>% mutate(SHIFT_16_20 = rep_share_20 - rep_share_16)
test_shift_vals  <- test_join  %>% mutate(SHIFT_20_24 = rep_share_24 - rep_share_20)

summary(train_shift_vals$SHIFT_16_20)
summary(test_shift_vals$SHIFT_20_24)

sd(train_shift_vals$SHIFT_16_20, na.rm=TRUE)
sd(test_shift_vals$SHIFT_20_24, na.rm=TRUE)


shift <- test_shift_vals$SHIFT_20_24
quantile(shift, c(.001,.01,.05,.5,.95,.99,.999), na.rm=TRUE)
```

```{r}
library(tidyverse)

# --- Load 2020 + 2024 with totalvotes kept ---
df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv") %>%
  select(county_fips, county_name, state_po,
         rep_share_20 = rep_share,
         totalvotes_20 = total_votes)

df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv") %>%
  select(county_fips,
         rep_share_24 = rep_share,
         totalvotes_24 = total_votes)

# --- Join + compute shift ---
shift_tbl <- inner_join(df_20, df_24, by = "county_fips") %>%
  mutate(
    SHIFT_20_24 = rep_share_24 - rep_share_20,
    votes_min   = pmin(totalvotes_20, totalvotes_24),
    votes_max   = pmax(totalvotes_20, totalvotes_24)
  )

# --- Bottom 25 (biggest negative shifts) ---
bottom25 <- shift_tbl %>%
  arrange(SHIFT_20_24) %>%
  slice(1:25) %>%
  select(county_fips, county_name, state_po,
         rep_share_20, rep_share_24, SHIFT_20_24,
         totalvotes_20, totalvotes_24, votes_min, votes_max)

# --- Top 25 (biggest positive shifts) ---
top25 <- shift_tbl %>%
  arrange(desc(SHIFT_20_24)) %>%
  slice(1:25) %>%
  select(county_fips, county_name, state_po,
         rep_share_20, rep_share_24, SHIFT_20_24,
         totalvotes_20, totalvotes_24, votes_min, votes_max)

cat("\nBottom 25 shifts (most negative):\n")
print(bottom25)

cat("\nTop 25 shifts (most positive):\n")
print(top25)

```

```{r}

library(dplyr)

var = 20207
og <- read.csv("data/prep/counties/countypres_2000-2024.csv")

# Your original check (still fine)
og[og$county_fips == var, c("county_fips", "party", "year", "candidatevotes", "totalvotes", "mode")]


shift_tbl[shift_tbl$county_fips == var, ]
```

```{r}
library(dplyr)
library(stringr)

og <- read.csv("data/prep/counties/countypres_2000-2024.csv")

pres_2024 <- og %>%
  filter(year == 2024) %>%
  mutate(
    office_up = toupper(trimws(as.character(office))),
    candidate = str_squish(as.character(candidate)),
    party     = str_squish(as.character(party)),
    cand_up   = toupper(candidate),
    party_up  = toupper(party)
  ) %>%
  filter(grepl("PRESIDENT", office_up))

# define the same "summary row" flag
pres_2024 <- pres_2024 %>%
  mutate(
    is_summary_row =
      grepl("TOTAL", cand_up) |
      grepl("VOTES CAST", cand_up) |
      ((party_up == "" | is.na(party_up)) &
         !is.na(candidatevotes) & !is.na(totalvotes) &
         candidatevotes == totalvotes)
  )

# For each county_fips: do we have ANY non-summary candidate rows?
county_summary_status <- pres_2024 %>%
  group_by(county_fips) %>%
  summarise(
    n_rows = n(),
    n_non_summary = sum(!is_summary_row, na.rm = TRUE),
    .groups = "drop"
  )

# counties that would be fully dropped by summary-row removal
all_summary_counties <- county_summary_status %>%
  filter(n_non_summary == 0)

cat("2024 counties with ONLY summary rows:", nrow(all_summary_counties), "\n")
head(all_summary_counties, 20)
```
