---
title: "county_models"
format: html
editor: visual
---

## 5-Fold CV for 2016 (Predicting Trump Share)

```{r}

library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2016
clean_16 <- read_csv("data/prep/counties/clean_county_emb_16_17.csv") 

# Select features and remove rows with missing values
data_rf <- clean_16 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean") # Only embedding means
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

write_csv(full_results, "data/results/county16_m1_results.csv")

```

## 5-Fold CV for 2020 (Predicting Trump Share)

This script takes ~20 sec to run

```{r}

library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2020
clean_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv") 

# Select features and remove rows with missing values
data_rf <- clean_20 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean") # Only embedding means
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

write_csv(full_results, "data/results/county20_m1_results.csv")

```
R-Squared:  0.647
RMSE:       0.0958
Mean Abs Error: 0.0709


## 5-Fold CV for 2024 (Predicting Trump Share)

This script takes ~20 sec to run

```{r}

library(tidyverse)
library(ranger)
library(scales) # for comma labels in plots

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2024
clean_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv") 

# Select features and remove rows with missing values
data_rf <- clean_24 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean") # Only embedding means
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

# Save results 
write_csv(full_results, "data/results/county24_m1_results.csv")

```
R-Squared:  0.632
RMSE:       0.1189
Mean Abs Error: 0.094

## 5-FOLD CV, demographics 2020
```{r}

library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2020
clean_20 <- read_csv("data/prep/counties/master_county_emb_20.csv") 

# Select features and remove rows with missing values
data_rf <- clean_20 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean"), # Only embedding means
    pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, # Education
    unemployment_rate, civilian_labor_force # Labor
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

write_csv(full_results, "data/results/county20_m2_results.csv")

```
R-Squared:  0.742
RMSE:       0.0819
Mean Abs Error: 0.0596

## 5-Fold CV, demographics 2024
```{r}


library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2024
clean_24 <- read_csv("data/prep/counties/master_county_emb_24.csv") 

# Select features and remove rows with missing values
data_rf <- clean_24 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    ends_with("_mean"), # Only embedding means
    pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, # Education
    unemployment_rate, civilian_labor_force # Labor
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - vote_density - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

write_csv(full_results, "data/results/county24_m2_results.csv")

```
R-Squared:  0.669
RMSE:       0.1127
Mean Abs Error: 0.0881


```{r}
summary(lm(rep_share ~ pct_less_than_hs + pct_hs_only + pct_some_college + pct_bachelors_plus + unemployment_rate + civilian_labor_force, data = clean_20))

summary(lm(rep_share ~ pct_less_than_hs + pct_hs_only + pct_some_college + pct_bachelors_plus + unemployment_rate + civilian_labor_force, data = clean_24))
```


### Visualizing results for county models

```{r}
# ==========================================
# 4. Overall Performance Metrics
# ==========================================

#full_results <- read.csv("data/results/county20_m1_results.csv")
full_results <- read.csv("data/results/county24_m2_results.csv")

# Calculate metrics on the full dataset (all folds combined)
rss  <- sum(full_results$Error^2)
tss  <- sum((full_results$TRUMPSHARE - mean(full_results$TRUMPSHARE))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(full_results$Error^2))

cat("\n==========================================")
cat("\nOVERALL 5-FOLD CV RESULTS")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4))
cat("\nMean Abs Error:", round(mean(full_results$Abs_Error), 4))
cat("\n==========================================\n")

# ==========================================
# 5. Plot 1: Prediction Error vs. Density
# ==========================================

plot1 <- ggplot(full_results, aes(x = vote_density, y = Error)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  scale_x_log10(labels = scales::comma) + 
  labs(
    title = "Prediction Error vs. Voter Density (5-Fold CV)",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 3)),
    x = "Voters per km² (Log Scale)",
    y = "Error (Predicted - Actual)"
  ) +
  theme_minimal()

print(plot1)

# ==========================================
# 6. Plot 2: R-squared per Density Decile
# ==========================================

decile_stats <- full_results %>%
  mutate(density_decile = ntile(vote_density, 10)) %>%
  group_by(density_decile) %>%
  summarize(
    avg_density = median(vote_density),
    n = n(),
    RSS = sum(Error^2),
    TSS = sum((TRUMPSHARE - mean(TRUMPSHARE))^2),
    # Safety check: avoid dividing by zero if TSS is 0
    R2 = if_else(TSS > 1e-6, 1 - (RSS / TSS), 0)
  )

plot2 <- ggplot(decile_stats, aes(x = factor(density_decile), y = R2)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(R2, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance by Density Decile",
    subtitle = "Decile 1 = Most Rural, Decile 10 = Most Urban",
    x = "Density Decile",
    y = "R-squared"
  ) +
  theme_minimal()

print(plot2)

# ==========================================
# 7. Plot 3: Predicted vs. Actual Trump Share
# ==========================================

plot3 <- ggplot(full_results, aes(x = TRUMPSHARE, y = Predicted_Share)) +
  # Add points with slight transparency
  geom_point(alpha = 0.4, size = 1, color = "darkslategrey") +
  
  # The 'Perfect Prediction' line (y = x)
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  
  # Add a trend line to see if the model is biased
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  
  # Format axes as percentages for readability
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  
  labs(
    title = "Actual vs. Predicted Trump Vote Share",
    subtitle = paste0("County-Level 5-Fold CV | R² = ", round(r2, 3)),
    x = "Actual Trump Share",
    y = "Predicted Trump Share",
    caption = "Red dashed line represents perfect prediction."
  ) +
  theme_minimal()

print(plot3)
```


```{r}

# Compare the "Spread" of the target variable
cat("2020 SD:", sd(clean_20$rep_share, na.rm=TRUE), "\n")
cat("2024 SD:", sd(clean_24$rep_share, na.rm=TRUE), "\n")

```


## 5-Fold CV for CHANGE in Trumpshare, 2024 change from 2020

This script takes ~20 sec to run

```{r}

# ==========================================
# 1. Load & Join Data (2020 vs 2024)
# ==========================================

# Load clean datasets
# Note: Added `total_votes_20 = total_votes` assuming your CSV has a total_votes column!
df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv") %>%
  select(county_fips, rep_share_20 = rep_share, total_votes_20 = total_votes)

df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv") %>%
  # We rename the 2024 share and total votes
  rename(rep_share_24 = rep_share, total_votes_24 = total_votes)

# Join the years to calculate the shift
data_shift <- inner_join(df_20, df_24, by = "county_fips") %>%
  mutate(
    # TARGET VARIABLE: Positive = Shift toward Trump, Negative = Shift toward Harris/Biden
    SHIFT_20_24 = rep_share_24 - rep_share_20,
    # NEW: Calculate the percentage change in total votes
    VOTE_CHANGE = (total_votes_24 - total_votes_20) / total_votes_20
  ) %>%
  # Prepare for modeling: Keep ID, Shift, 2020 Baseline, Vote Change, and Embeddings
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_20,    # Kept for analysis/plotting, NOT for training
    SHIFT_20_24,     # The Target
    VOTE_CHANGE,     # Kept temporarily for filtering
    ends_with("_mean") # The Predictors (2024 embeddings)
  ) %>%
  na.omit()

cat("Data Loaded. Analyzing shift for", nrow(data_shift), "counties.\n")
cat("Average Shift:", round(mean(data_shift$SHIFT_20_24) * 100, 2), "points\n")

# ==========================================
# FILTER: Remove extreme turnout changes using IQR (2x)
# ==========================================

# 1. Calculate IQR bounds for 2020-2024 Vote Change
q1 <- quantile(data_shift$VOTE_CHANGE, 0.25, na.rm = TRUE)
q3 <- quantile(data_shift$VOTE_CHANGE, 0.75, na.rm = TRUE)
iqr <- q3 - q1
lower_bound <- q1 - 2 * iqr
upper_bound <- q3 + 2 * iqr

# 2. Plot a scatterplot of change in voteshare vs change in total_votes
p_outliers <- ggplot(data_shift, aes(x = VOTE_CHANGE, y = SHIFT_20_24)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  # Add red lines to show exactly where we are cutting the data dynamically
  geom_vline(xintercept = c(lower_bound, upper_bound), linetype = "dashed", color = "red") +
  labs(
    title = "Vote Share Shift vs. Total Votes Change (2020-2024)",
    subtitle = sprintf("Red lines = 2*IQR bounds (%.1f%% to %.1f%%)", lower_bound*100, upper_bound*100),
    x = "Change in Total Votes (2020 to 2024)",
    y = "Shift in GOP Vote Share"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p_outliers)

cat("\n[Turnout IQR filter] Before:", nrow(data_shift), "counties\n")

# 3. Filter out counties outside the IQR bounds
data_filtered <- data_shift %>%
  filter(VOTE_CHANGE >= lower_bound & VOTE_CHANGE <= upper_bound)

# Print how many counties were removed
counties_removed <- nrow(data_shift) - nrow(data_filtered)

cat("[Turnout IQR filter] After: ", nrow(data_filtered), "counties\n")
cat("[Turnout IQR filter] Removed:", counties_removed, "counties due to extreme total vote swing\n")


# Create a dataframe of just the dropped counties
dropped_counties <- data_shift %>% # (assuming you have the pre-filter data)
  filter(VOTE_CHANGE < lower_bound | VOTE_CHANGE > upper_bound)

# Look at which states lost the most counties
table(dropped_counties$state_po) %>% sort(decreasing = TRUE) %>% head(10)

# 4. Reassign back to data_shift and drop VOTE_CHANGE so it doesn't leak into the Random Forest model
data_shift <- data_filtered %>% 
  select(-VOTE_CHANGE)

# Optional quick summary of remaining shift range
cat("[Turnout IQR filter] Vote Share Shift range now:",
    round(min(data_shift$SHIFT_20_24), 4), "to",
    round(max(data_shift$SHIFT_20_24), 4), "\n")


# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(55) # New seed for luck
data_shift <- data_shift %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

all_preds <- list()
importances <- list() # Store variable importance for each fold

cat("\nStarting CV to predict Swing/Shift...\n")

# ==========================================
# 3. CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split
  train_df <- data_shift %>% filter(fold_id != k)
  test_df  <- data_shift %>% filter(fold_id == k)
  
  # B. Train Model
  # Predict SHIFT using only Embeddings
  # Note: We exclude rep_share_20 to see if *Geography alone* predicts the swing
  rf_model <- ranger(
    SHIFT_20_24 ~ . - county_fips - county_name - state_po - vote_density - rep_share_20 - fold_id,
    data       = train_df,
    num.trees  = 500,
    mtry       = 21, # ~64 predictors / 3
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Predictions
  test_df$Predicted_Shift <- preds
  all_preds[[k]] <- test_df
  
  # E. Store Importance (optional, for curiosity)
  importances[[k]] <- rf_model$variable.importance
  
  cat("  Fold", k, "done.\n")
}

# Combine Results
results <- bind_rows(all_preds) %>%
  mutate(
    Error = Predicted_Shift - SHIFT_20_24,
    Abs_Error = abs(Error)
  )

# ==========================================
# 4. Metrics
# ==========================================

rss  <- sum(results$Error^2)
tss  <- sum((results$SHIFT_20_24 - mean(results$SHIFT_20_24))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(results$Error^2))

cat("\n==========================================")
cat("\nMODEL RESULTS: PREDICTING THE SWING")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4), "(", round(rmse*100, 2), "percentage points )")
cat("\n==========================================\n")

# ==========================================
# 5. Visualization 1: Actual vs. Predicted Shift
# ==========================================

# A 45-degree line plot to see if the model captures the direction of the shift
p1 <- ggplot(results, aes(x = SHIFT_20_24, y = Predicted_Shift)) +
  geom_point(alpha = 0.3, size = 1, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Can Satellites Predict Political Swings?",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 4)),
    x = "Actual Shift (2020 -> 2024)",
    y = "Predicted Shift (Based on Embeddings)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p1)

# ==========================================
# 6. Visualization 2: Error Bias by Baseline Partisanship
# ==========================================

# Did the model fail more in Deep Red or Deep Blue counties?
# We bin the 2020 Vote Share to see where the "surprises" happened.

p2 <- ggplot(results, aes(x = rep_share_20, y = Error)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", color = "blue", fill = "gray90") +
  labs(
    title = "Where were the surprises?",
    subtitle = "Positive Error = Model expected a more pro-Trump shift than happened",
    x = "2020 GOP Vote Share (Baseline)",
    y = "Prediction Error (Pred Swing - Actual Swing)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal()

print(p2)

# ==========================================
# 7. Visualization 3: Top Predictive Features
# ==========================================

# Average importance across folds
avg_imp <- bind_rows(lapply(importances, function(x) as.data.frame(t(x)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Importance") %>%
  arrange(desc(Importance)) %>%
  slice(1:15)

p3 <- ggplot(avg_imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Which Embeddings Predict the Swing?",
    subtitle = "Top 15 Features driving the change from 2020 to 2024",
    x = "",
    y = "Variable Importance (Impurity)"
  ) +
  theme_minimal()

print(p3)

# ==========================================
# 8. Visualization 4: R² by Vote Density Decile
# ==========================================

# Helper to compute R² safely (returns NA if variance is zero)
r2_fun <- function(actual, pred) {
  rss <- sum((pred - actual)^2, na.rm = TRUE)
  tss <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  if (is.na(tss) || tss == 0) return(NA_real_)
  1 - rss / tss
}

# Create density deciles + compute per-decile metrics
r2_by_decile <- results %>%
  mutate(
    density_decile = ntile(vote_density, 10)
  ) %>%
  group_by(density_decile) %>%
  summarise(
    n = n(),
    r2 = r2_fun(SHIFT_20_24, Predicted_Shift),
    rmse = sqrt(mean((Predicted_Shift - SHIFT_20_24)^2, na.rm = TRUE)),
    density_min = min(vote_density, na.rm = TRUE),
    density_max = max(vote_density, na.rm = TRUE),
    .groups = "drop"
  )

print(r2_by_decile)

# Bar plot of R² by decile
p4 <- ggplot(r2_by_decile, aes(x = factor(density_decile), y = r2)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Model Performance by Vote Density",
    subtitle = "Out-of-sample R² computed within each vote_density decile",
    x = "Vote Density Decile (1 = lowest density, 10 = highest)",
    y = "R² (within decile)"
  ) +
  theme_minimal()

print(p4)

write_csv(results, "data/results/county_change_m1_results.csv")


```


## Predicting only with the demographics

```{r}


library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Clean Data
# ==========================================

# 2020
clean_20 <- read_csv("data/prep/counties/master_county_emb_20.csv") 

# Select features and remove rows with missing values
data_rf <- clean_20 %>%
  select(
    county_fips, 
    vote_density, 
    TRUMPSHARE = rep_share, 
    #ends_with("_mean"), # Only embedding means
    pct_less_than_hs, pct_hs_only, pct_some_college, pct_bachelors_plus, # Education
    unemployment_rate, civilian_labor_force # Labor
  ) %>%
  na.omit()

# ==========================================
# 2. Setup 5-Fold Cross-Validation
# ==========================================

set.seed(10)

# Create 5 random folds
# We shuffle the data first, then assign a fold number (1 through 5)
data_rf <- data_rf %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

# Initialize an empty list to store predictions
all_predictions <- list()

cat("Starting 5-Fold Cross-Validation on", nrow(data_rf), "counties...\n")

# ==========================================
# 3. The CV Loop
# ==========================================

for(k in 1:5) {
  
  # A. Split Data
  train_df <- data_rf %>% filter(fold_id != k)
  test_df  <- data_rf %>% filter(fold_id == k)
  
  # B. Train Model
  # mtry strategy: Number of predictors / 3 is standard for regression
  n_features <- ncol(train_df) - 4 # Subtract ID, density, share, fold_id
  my_mtry    <- floor(n_features / 3)
  
  rf_model <- ranger(
    TRUMPSHARE ~ . - county_fips - fold_id, 
    data       = train_df,
    num.trees  = 500,
    mtry       = my_mtry, # Dynamic calculation (approx 21)
    importance = "impurity",
    seed       = 42
  )
  
  # C. Predict
  preds <- predict(rf_model, data = test_df)$predictions
  
  # D. Store Results
  # We add the predictions back to the test dataframe
  test_df$Predicted_Share <- preds
  all_predictions[[k]] <- test_df
  
  cat("  Fold", k, "complete. Rows predicted:", nrow(test_df), "\n")
}

# Combine all 5 folds back into one big dataframe
full_results <- bind_rows(all_predictions) %>%
  mutate(
    Error = Predicted_Share - TRUMPSHARE,
    Abs_Error = abs(Error)
  )

# ==========================================
# 4. Overall Performance Metrics
# ==========================================

#full_results <- read.csv("data/results/county20_m1_results.csv")
#full_results <- read.csv("data/results/county24_m2_results.csv")

# Calculate metrics on the full dataset (all folds combined)
rss  <- sum(full_results$Error^2)
tss  <- sum((full_results$TRUMPSHARE - mean(full_results$TRUMPSHARE))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(full_results$Error^2))

cat("\n==========================================")
cat("\nOVERALL 5-FOLD CV RESULTS")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4))
cat("\nMean Abs Error:", round(mean(full_results$Abs_Error), 4))
cat("\n==========================================\n")

# ==========================================
# 5. Plot 1: Prediction Error vs. Density
# ==========================================

plot1 <- ggplot(full_results, aes(x = vote_density, y = Error)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  scale_x_log10(labels = scales::comma) + 
  labs(
    title = "Prediction Error vs. Voter Density (5-Fold CV)",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 3)),
    x = "Voters per km² (Log Scale)",
    y = "Error (Predicted - Actual)"
  ) +
  theme_minimal()

print(plot1)

# ==========================================
# 6. Plot 2: R-squared per Density Decile
# ==========================================

decile_stats <- full_results %>%
  mutate(density_decile = ntile(vote_density, 10)) %>%
  group_by(density_decile) %>%
  summarize(
    avg_density = median(vote_density),
    n = n(),
    RSS = sum(Error^2),
    TSS = sum((TRUMPSHARE - mean(TRUMPSHARE))^2),
    # Safety check: avoid dividing by zero if TSS is 0
    R2 = if_else(TSS > 1e-6, 1 - (RSS / TSS), 0)
  )

plot2 <- ggplot(decile_stats, aes(x = factor(density_decile), y = R2)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(R2, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance by Density Decile",
    subtitle = "Decile 1 = Most Rural, Decile 10 = Most Urban",
    x = "Density Decile",
    y = "R-squared"
  ) +
  theme_minimal()

print(plot2)

# ==========================================
# 7. Plot 3: Predicted vs. Actual Trump Share
# ==========================================

plot3 <- ggplot(full_results, aes(x = TRUMPSHARE, y = Predicted_Share)) +
  # Add points with slight transparency
  geom_point(alpha = 0.4, size = 1, color = "darkslategrey") +
  
  # The 'Perfect Prediction' line (y = x)
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  
  # Add a trend line to see if the model is biased
  geom_smooth(method = "lm", color = "blue", se = TRUE) +
  
  # Format axes as percentages for readability
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  
  labs(
    title = "Actual vs. Predicted Trump Vote Share",
    subtitle = paste0("County-Level 5-Fold CV | R² = ", round(r2, 3)),
    x = "Actual Trump Share",
    y = "Predicted Trump Share",
    caption = "Red dashed line represents perfect prediction."
  ) +
  theme_minimal()

print(plot3)

```

train on 2016 -> 2020, then predict 2020 -> 2024 shifts, WITHOUT training on 2020 -> 2024.

```{r}

library(tidyverse)
library(ranger)

# ==============================================================================
# GOAL:
#   Train on 2016 -> 2020 shifts (using 2020 embeddings as predictors),
#   then predict 2020 -> 2024 shifts (using 2024 embeddings as predictors),
#   WITHOUT training on 2020 -> 2024.
# ==============================================================================


# ==========================================
# 1. Load & Join Data for TRAIN (2016 vs 2020)
# ==========================================

df_16 <- read_csv("data/prep/counties/clean_county_emb_16_17.csv") %>%
  select(county_fips, rep_share_16 = rep_share, total_votes_16 = total_votes)

df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv") %>%
  rename(rep_share_20 = rep_share, total_votes_20 = total_votes)

# TRAIN set: 2016 -> 2020 shift, predictors = 2020 embeddings
train_shift <- inner_join(df_16, df_20, by = "county_fips") %>%
  mutate(
    SHIFT_16_20 = rep_share_20 - rep_share_16,
    VOTE_CHANGE_16_20 = (total_votes_20 - total_votes_16) / total_votes_16
  ) %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_16,
    rep_share_20,      # kept for analysis, not used for training
    SHIFT_16_20,       # target for training
    VOTE_CHANGE_16_20, # kept temporarily for filtering
    ends_with("_mean") # predictors = 2020 embeddings
  ) %>%
  na.omit()

cat("TRAIN data loaded (2016->2020) for", nrow(train_shift), "counties.\n")
cat("Avg TRAIN shift:", round(mean(train_shift$SHIFT_16_20) * 100, 2), "points\n")


# ==========================================
# 2. Load & Join Data for TEST (2020 vs 2024)
# ==========================================

df_20_base <- read_csv("data/prep/counties/clean_county_emb_20.csv") %>%
  select(county_fips, rep_share_20 = rep_share, total_votes_20 = total_votes)

df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv") %>%
  rename(rep_share_24 = rep_share, total_votes_24 = total_votes)

# TEST set: 2020 -> 2024 shift, predictors = 2024 embeddings
test_shift <- inner_join(df_20_base, df_24, by = "county_fips") %>%
  mutate(
    SHIFT_20_24 = rep_share_24 - rep_share_20,
    VOTE_CHANGE_20_24 = (total_votes_24 - total_votes_20) / total_votes_20
  ) %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_20,      # kept for analysis, not used for training
    SHIFT_20_24,       # target only for evaluation
    VOTE_CHANGE_20_24, # kept temporarily for filtering
    ends_with("_mean") # predictors = 2024 embeddings
  ) %>%
  na.omit()

cat("\nTEST data loaded (2020->2024) for", nrow(test_shift), "counties.\n")
cat("Avg TEST shift:", round(mean(test_shift$SHIFT_20_24) * 100, 2), "points\n")

# ==========================================
# FILTER: Remove extreme turnout changes using IQR (2x)
# ==========================================

# --- TRAIN outliers (2016 -> 2020) ---
# 1. Calculate IQR bounds for TRAIN
q1_train <- quantile(train_shift$VOTE_CHANGE_16_20, 0.25, na.rm = TRUE)
q3_train <- quantile(train_shift$VOTE_CHANGE_16_20, 0.75, na.rm = TRUE)
iqr_train <- q3_train - q1_train
lower_train <- q1_train - 2 * iqr_train
upper_train <- q3_train + 2 * iqr_train

# 2. Plot TRAIN with dynamic bounds
p_outliers_train <- ggplot(train_shift, aes(x = VOTE_CHANGE_16_20, y = SHIFT_16_20)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_vline(xintercept = c(lower_train, upper_train), linetype = "dashed", color = "red") +
  labs(
    title = "TRAIN: Vote Share Shift vs. Total Votes Change (2016-2020)",
    subtitle = sprintf("Red lines = 2*IQR bounds (%.1f%% to %.1f%%)", lower_train*100, upper_train*100),
    x = "Change in Total Votes (2016 to 2020)",
    y = "Shift in GOP Vote Share"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p_outliers_train)

# 3. Filter TRAIN
n_train_before <- nrow(train_shift)
train_shift <- train_shift %>% 
  filter(VOTE_CHANGE_16_20 >= lower_train & VOTE_CHANGE_16_20 <= upper_train)
n_train_after <- nrow(train_shift)

cat("\n[Turnout IQR filter | TRAIN] Before:", n_train_before,
    "| After:", n_train_after,
    "| Removed:", (n_train_before - n_train_after), "counties\n")

# Drop temporary column
train_shift <- train_shift %>% select(-VOTE_CHANGE_16_20)


# --- TEST outliers (2020 -> 2024) ---
# 1. Calculate IQR bounds for TEST
q1_test <- quantile(test_shift$VOTE_CHANGE_20_24, 0.25, na.rm = TRUE)
q3_test <- quantile(test_shift$VOTE_CHANGE_20_24, 0.75, na.rm = TRUE)
iqr_test <- q3_test - q1_test
lower_test <- q1_test - 2 * iqr_test
upper_test <- q3_test + 2 * iqr_test

# 2. Plot TEST with dynamic bounds
p_outliers_test <- ggplot(test_shift, aes(x = VOTE_CHANGE_20_24, y = SHIFT_20_24)) +
  geom_point(alpha = 0.5, color = "darkorange") +
  geom_vline(xintercept = c(lower_test, upper_test), linetype = "dashed", color = "red") +
  labs(
    title = "TEST: Vote Share Shift vs. Total Votes Change (2020-2024)",
    subtitle = sprintf("Red lines = 2*IQR bounds (%.1f%% to %.1f%%)", lower_test*100, upper_test*100),
    x = "Change in Total Votes (2020 to 2024)",
    y = "Shift in GOP Vote Share"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p_outliers_test)

# 3. Filter TEST
n_test_before <- nrow(test_shift)
test_shift <- test_shift %>% 
  filter(VOTE_CHANGE_20_24 >= lower_test & VOTE_CHANGE_20_24 <= upper_test)
n_test_after <- nrow(test_shift)

cat("\n[Turnout IQR filter | TEST]  Before:", n_test_before,
    "| After:", n_test_after,
    "| Removed:", (n_test_before - n_test_after), "counties\n")

# Drop temporary column
test_shift <- test_shift %>% select(-VOTE_CHANGE_20_24)

# ==========================================
# 3. Make sure TRAIN and TEST have identical predictor columns
# ==========================================

id_cols_train <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_16", "rep_share_20", "SHIFT_16_20")
id_cols_test  <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_20", "SHIFT_20_24")

x_cols_train <- setdiff(names(train_shift), id_cols_train)
x_cols_test  <- setdiff(names(test_shift),  id_cols_test)

common_x <- intersect(x_cols_train, x_cols_test)

# Keep only the shared predictors (safe if columns differ across years)
train_shift <- train_shift %>% select(all_of(id_cols_train), all_of(common_x))
test_shift  <- test_shift  %>% select(all_of(id_cols_test),  all_of(common_x))

cat("\nUsing", length(common_x), "shared embedding predictors.\n")


# ==========================================
# 4. 5-Fold CV ON TRAIN (2016->2020), predict TEST (2020->2024)
# ==========================================

set.seed(55)
train_shift <- train_shift %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

test_preds_by_fold <- list()
importances <- list()

cat("\nStarting CV: train on 2016->2020, predict 2020->2024...\n")

for (k in 1:5) {

  # A. Train split
  train_df <- train_shift %>% filter(fold_id != k)

  # B. Fit model on 2016->2020 shift
  rf_model <- ranger(
    SHIFT_16_20 ~ . - county_fips - county_name - state_po - vote_density - rep_share_16 - rep_share_20 - fold_id,
    data       = train_df,
    num.trees  = 500,
    mtry       = 21,           # keep your setting; tweak if you want
    importance = "impurity",
    seed       = 42
  )

  # C. Predict 2020->2024 shifts for ALL test counties using this fold's model
  preds <- predict(rf_model, data = test_shift)$predictions

  tmp <- test_shift %>%
    mutate(
      fold_model = k,
      Predicted_Shift = preds
    )

  test_preds_by_fold[[k]] <- tmp
  importances[[k]] <- rf_model$variable.importance

  cat("  Fold", k, "model trained (on 2016->2020) and applied to 2020->2024.\n")
}

# Average predictions across the 5 independently-trained models
results <- bind_rows(test_preds_by_fold) %>%
  group_by(county_fips, county_name, state_po, vote_density, rep_share_20, SHIFT_20_24) %>%
  summarise(
    Predicted_Shift = mean(Predicted_Shift),
    .groups = "drop"
  ) %>%
  mutate(
    Error = Predicted_Shift - SHIFT_20_24,
    Abs_Error = abs(Error)
  )


# ==========================================
# 5. Metrics (evaluated on 2020->2024)
# ==========================================

rss  <- sum(results$Error^2)
tss  <- sum((results$SHIFT_20_24 - mean(results$SHIFT_20_24))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(results$Error^2))

cat("\n==========================================")
cat("\nTRANSFER MODEL RESULTS")
cat("\nTrain: 2016->2020 | Test: 2020->2024")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4), "(", round(rmse*100, 2), "percentage points )")
cat("\n==========================================\n")


# ==========================================
# 6. Visualization 1: Actual vs. Predicted Shift
# ==========================================

p1 <- ggplot(results, aes(x = SHIFT_20_24, y = Predicted_Shift)) +
  geom_point(alpha = 0.3, size = 1, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Out-of-period generalization: 2016→2020 trained, 2020→2024 predicted",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 4)),
    x = "Actual Shift (2020 → 2024)",
    y = "Predicted Shift (trained on 2016 → 2020)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p1)


# ==========================================
# 7. Visualization 2: Error Bias by Baseline Partisanship
# ==========================================

p2 <- ggplot(results, aes(x = rep_share_20, y = Error)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", color = "blue", fill = "gray90") +
  labs(
    title = "Where were the surprises (transfer model)?",
    subtitle = "Positive Error = model expected a more pro-GOP shift than happened",
    x = "2020 GOP Vote Share (Baseline)",
    y = "Prediction Error (Pred - Actual)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal()

print(p2)


# ==========================================
# 8. Visualization 3: Top Predictive Features (avg across folds)
# ==========================================

avg_imp <- bind_rows(lapply(importances, function(x) as.data.frame(t(x)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Importance") %>%
  arrange(desc(Importance)) %>%
  slice(1:15)

p3 <- ggplot(avg_imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Which embeddings predict swings (learned from 2016→2020)?",
    subtitle = "Top 15 features by impurity importance (avg across folds)",
    x = "",
    y = "Variable Importance (Impurity)"
  ) +
  theme_minimal()

print(p3)


# ==========================================
# 9. Visualization 4: R² by Vote Density Decile (TRANSFER MODEL)
#   Drop this right after p3 (and before write_csv).
# ==========================================

# Helper to compute R² safely (returns NA if variance is zero)
r2_fun <- function(actual, pred) {
  rss <- sum((pred - actual)^2, na.rm = TRUE)
  tss <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  if (is.na(tss) || tss == 0) return(NA_real_)
  1 - rss / tss
}

r2_by_density_decile <- results %>%
  mutate(
    density_decile = ntile(vote_density, 10)
    # If density is super skewed, you may prefer:
    # density_decile = ntile(log1p(vote_density), 10)
  ) %>%
  group_by(density_decile) %>%
  summarise(
    n = n(),
    r2 = r2_fun(SHIFT_20_24, Predicted_Shift),
    rmse = sqrt(mean((Predicted_Shift - SHIFT_20_24)^2, na.rm = TRUE)),
    density_min = min(vote_density, na.rm = TRUE),
    density_max = max(vote_density, na.rm = TRUE),
    .groups = "drop"
  )

print(r2_by_density_decile)

p4 <- ggplot(r2_by_density_decile, aes(x = factor(density_decile), y = r2)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Transfer-model performance by vote density",
    subtitle = "R² computed within each vote_density decile (test period: 2020→2024)",
    x = "Vote Density Decile (1 = lowest density, 10 = highest)",
    y = "R² (within decile)"
  ) +
  theme_minimal()

print(p4)

# ==========================================
# 9. Save
# ==========================================

write_csv(results, "data/results/county_change_m2_results.csv")
cat("Saved")



```


```{r}

## CHANGE to voter turnout outliers

library(tidyverse)
library(ranger)

# ==============================================================================
# GOAL (EARLY-YEAR EMBEDDINGS):
#   Train on 2016 -> 2020 shifts using **2016 embeddings** as predictors,
#   then predict 2020 -> 2024 shifts using **2020 embeddings** as predictors,
#   WITHOUT training on 2020 -> 2024.
# ==============================================================================


# ==========================================
# 1. Load Data
# ==========================================

# Full files (keep embeddings/metadata)
df_16_full <- read_csv("data/prep/counties/clean_county_emb_16_17.csv")
df_20_full <- read_csv("data/prep/counties/clean_county_emb_20.csv")
df_24_full <- read_csv("data/prep/counties/clean_county_emb_24.csv")

# Shares-only extracts (to avoid duplicate cols in joins)
df_16_share <- df_16_full %>% select(county_fips, rep_share_16 = rep_share)
df_20_share <- df_20_full %>% select(county_fips, rep_share_20 = rep_share)
df_24_share <- df_24_full %>% select(county_fips, rep_share_24 = rep_share)


# ==========================================
# 2. TRAIN: 2016 -> 2020 shift, predictors = 2016 embeddings
# ==========================================

train_shift <- inner_join(df_16_share, df_20_share, by = "county_fips") %>%
  # attach 2016 embeddings + metadata we want to keep
  inner_join(
    df_16_full %>%
      select(county_fips, county_name, state_po, vote_density, ends_with("_mean")),
    by = "county_fips"
  ) %>%
  mutate(
    SHIFT_16_20 = rep_share_20 - rep_share_16
  ) %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_16,
    rep_share_20,      # kept for analysis, not used for training
    SHIFT_16_20,       # target for training
    ends_with("_mean") # predictors = 2016 embeddings
  ) %>%
  na.omit()

cat("TRAIN data loaded (2016->2020) for", nrow(train_shift), "counties.\n")
cat("Avg TRAIN shift:", round(mean(train_shift$SHIFT_16_20) * 100, 2), "points\n")


# ==========================================
# 3. TEST: 2020 -> 2024 shift, predictors = 2020 embeddings
# ==========================================

test_shift <- inner_join(df_20_share, df_24_share, by = "county_fips") %>%
  # attach 2020 embeddings + metadata we want to keep
  inner_join(
    df_20_full %>%
      select(county_fips, county_name, state_po, vote_density, ends_with("_mean")),
    by = "county_fips"
  ) %>%
  mutate(
    SHIFT_20_24 = rep_share_24 - rep_share_20
  ) %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_20,      # kept for analysis, not used for training
    SHIFT_20_24,       # target only for evaluation
    ends_with("_mean") # predictors = 2020 embeddings
  ) %>%
  na.omit()

cat("\nTEST data loaded (2020->2024) for", nrow(test_shift), "counties.\n")
cat("Avg TEST shift:", round(mean(test_shift$SHIFT_20_24) * 100, 2), "points\n")


# ==========================================
# OPTIONAL: Remove extreme shift outliers (|SHIFT| > 0.15)
#   Drop this RIGHT AFTER you create train_shift and test_shift (after na.omit()).
#   Easy to comment out as a block.
# ==========================================

outlier_cutoff <- 0.15

# --- TRAIN outliers (2016 -> 2020) ---
n_train_before <- nrow(train_shift)
train_shift <- train_shift %>% filter(abs(SHIFT_16_20) <= outlier_cutoff)
n_train_after <- nrow(train_shift)

cat("\n[Outlier filter | TRAIN] Before:", n_train_before,
    "| After:", n_train_after,
    "| Removed:", (n_train_before - n_train_after), "\n")
cat("[Outlier filter | TRAIN] Shift range now:",
    round(min(train_shift$SHIFT_16_20), 4), "to",
    round(max(train_shift$SHIFT_16_20), 4), "\n")

# --- TEST outliers (2020 -> 2024) ---
n_test_before <- nrow(test_shift)
test_shift <- test_shift %>% filter(abs(SHIFT_20_24) <= outlier_cutoff)
n_test_after <- nrow(test_shift)

cat("\n[Outlier filter | TEST]  Before:", n_test_before,
    "| After:", n_test_after,
    "| Removed:", (n_test_before - n_test_after), "\n")
cat("[Outlier filter | TEST]  Shift range now:",
    round(min(test_shift$SHIFT_20_24), 4), "to",
    round(max(test_shift$SHIFT_20_24), 4), "\n")


# ==========================================
# 4. Make sure TRAIN and TEST have identical predictor columns
# ==========================================

id_cols_train <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_16", "rep_share_20", "SHIFT_16_20")
id_cols_test  <- c("county_fips", "county_name", "state_po", "vote_density", "rep_share_20", "SHIFT_20_24")

x_cols_train <- setdiff(names(train_shift), id_cols_train)
x_cols_test  <- setdiff(names(test_shift),  id_cols_test)

common_x <- intersect(x_cols_train, x_cols_test)

train_shift <- train_shift %>% select(all_of(id_cols_train), all_of(common_x))
test_shift  <- test_shift  %>% select(all_of(id_cols_test),  all_of(common_x))

cat("\nUsing", length(common_x), "shared embedding predictors.\n")


# ==========================================
# 5. 5-Fold CV ON TRAIN (2016->2020), predict TEST (2020->2024)
# ==========================================

set.seed(55)
train_shift <- train_shift %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

test_preds_by_fold <- list()
importances <- list()

cat("\nStarting CV: train on 2016->2020 (2016 embeddings), predict 2020->2024 (2020 embeddings)...\n")

for (k in 1:5) {

  train_df <- train_shift %>% filter(fold_id != k)

  rf_model <- ranger(
    SHIFT_16_20 ~ . - county_fips - county_name - state_po - vote_density - rep_share_16 - rep_share_20 - fold_id,
    data       = train_df,
    num.trees  = 500,
    mtry       = 21,
    importance = "impurity",
    seed       = 42
  )

  preds <- predict(rf_model, data = test_shift)$predictions

  tmp <- test_shift %>%
    mutate(
      fold_model = k,
      Predicted_Shift = preds
    )

  test_preds_by_fold[[k]] <- tmp
  importances[[k]] <- rf_model$variable.importance

  cat("  Fold", k, "model trained (2016->2020) and applied to 2020->2024.\n")
}

results <- bind_rows(test_preds_by_fold) %>%
  group_by(county_fips, county_name, state_po, vote_density, rep_share_20, SHIFT_20_24) %>%
  summarise(
    Predicted_Shift = mean(Predicted_Shift),
    .groups = "drop"
  ) %>%
  mutate(
    Error = Predicted_Shift - SHIFT_20_24,
    Abs_Error = abs(Error)
  )


# ==========================================
# 6. Metrics (evaluated on 2020->2024)
# ==========================================

rss  <- sum(results$Error^2)
tss  <- sum((results$SHIFT_20_24 - mean(results$SHIFT_20_24))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(results$Error^2))

cat("\n==========================================")
cat("\nTRANSFER MODEL RESULTS (EARLY-YEAR EMBEDDINGS)")
cat("\nTrain: 2016->2020 (2016 emb) | Test: 2020->2024 (2020 emb)")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4), "(", round(rmse*100, 2), "percentage points )")
cat("\n==========================================\n")


# ==========================================
# 7. Visualizations (same as before)
# ==========================================

p1 <- ggplot(results, aes(x = SHIFT_20_24, y = Predicted_Shift)) +
  geom_point(alpha = 0.3, size = 1, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Early-year embeddings: 2016→2020 trained, 2020→2024 predicted",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 4)),
    x = "Actual Shift (2020 → 2024)",
    y = "Predicted Shift (trained on 2016 → 2020)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
print(p1)

p2 <- ggplot(results, aes(x = rep_share_20, y = Error)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", color = "blue", fill = "gray90") +
  labs(
    title = "Where were the surprises (transfer model, early embeddings)?",
    subtitle = "Positive Error = model expected a more pro-GOP shift than happened",
    x = "2020 GOP Vote Share (Baseline)",
    y = "Prediction Error (Pred - Actual)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal()
print(p2)

avg_imp <- bind_rows(lapply(importances, function(x) as.data.frame(t(x)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Importance") %>%
  arrange(desc(Importance)) %>%
  slice(1:15)

p3 <- ggplot(avg_imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Top features (learned from 2016 embeddings)",
    subtitle = "Top 15 features by impurity importance (avg across folds)",
    x = "",
    y = "Variable Importance (Impurity)"
  ) +
  theme_minimal()
print(p3)

# ==========================================
# 9. Visualization 4: R² by Vote Density Decile (TRANSFER MODEL)
#   Drop this right after p3 (and before write_csv).
# ==========================================

# Helper to compute R² safely (returns NA if variance is zero)
r2_fun <- function(actual, pred) {
  rss <- sum((pred - actual)^2, na.rm = TRUE)
  tss <- sum((actual - mean(actual, na.rm = TRUE))^2, na.rm = TRUE)
  if (is.na(tss) || tss == 0) return(NA_real_)
  1 - rss / tss
}

r2_by_density_decile <- results %>%
  mutate(
    density_decile = ntile(vote_density, 10)
    # If density is super skewed, you may prefer:
    # density_decile = ntile(log1p(vote_density), 10)
  ) %>%
  group_by(density_decile) %>%
  summarise(
    n = n(),
    r2 = r2_fun(SHIFT_20_24, Predicted_Shift),
    rmse = sqrt(mean((Predicted_Shift - SHIFT_20_24)^2, na.rm = TRUE)),
    density_min = min(vote_density, na.rm = TRUE),
    density_max = max(vote_density, na.rm = TRUE),
    .groups = "drop"
  )

print(r2_by_density_decile)

p4 <- ggplot(r2_by_density_decile, aes(x = factor(density_decile), y = r2)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Transfer-model performance by vote density",
    subtitle = "R² computed within each vote_density decile (test period: 2020→2024)",
    x = "Vote Density Decile (1 = lowest density, 10 = highest)",
    y = "R² (within decile)"
  ) +
  theme_minimal()

print(p4)

# ==========================================
# 8. Save
# ==========================================

write_csv(results, "data/results/county_change_m2_early_embeddings_results.csv")
cat("Saved: data/results/county_change_m2_early_embeddings_results.csv\n")
```


```{r}

## CHANGE to voter turnout outliers

library(tidyverse)
library(ranger)

# ==============================================================================
# GOAL (DELTA MODEL):
#   Train on 2016 -> 2020 shifts using DELTA embeddings (2020 - 2016),
#   then predict 2020 -> 2024 shifts using DELTA embeddings (2024 - 2020),
#   WITHOUT training on 2020 -> 2024 shifts.
#
#   Predictors: Axx_mean deltas only (suffix: _delta)
#   Target (train): SHIFT_16_20
#   Eval (test):    SHIFT_20_24
# ==============================================================================

# ==========================================
# 0. Load base files
# ==========================================

df_16 <- read_csv("data/prep/counties/clean_county_emb_16_17.csv", show_col_types = FALSE)
df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv",     show_col_types = FALSE)
df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv",     show_col_types = FALSE)

# We'll use only the _mean embeddings for deltas (consistent with your earlier scripts)
emb_cols_16 <- names(df_16) %>% keep(~ str_detect(.x, "_mean$"))
emb_cols_20 <- names(df_20) %>% keep(~ str_detect(.x, "_mean$"))
emb_cols_24 <- names(df_24) %>% keep(~ str_detect(.x, "_mean$"))

common_emb <- Reduce(intersect, list(emb_cols_16, emb_cols_20, emb_cols_24))

cat("Using", length(common_emb), "shared *_mean embedding columns for deltas.\n")

# ==========================================
# 1. TRAIN data: 2016 vs 2020 (delta = 2020 - 2016)
# ==========================================

df_16_small <- df_16 %>%
  select(
    county_fips,
    rep_share_16 = rep_share,
    all_of(common_emb)
  ) %>%
  rename_with(~ paste0(.x, "_16"), all_of(common_emb))

df_20_small <- df_20 %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_20 = rep_share,
    all_of(common_emb)
  ) %>%
  rename_with(~ paste0(.x, "_20"), all_of(common_emb))

train_join <- inner_join(df_16_small, df_20_small, by = "county_fips") %>%
  mutate(SHIFT_16_20 = rep_share_20 - rep_share_16)

# Build delta predictors safely (NO cur_column)
train_deltas <- purrr::map_dfc(common_emb, function(nm) {
  train_join[[paste0(nm, "_20")]] - train_join[[paste0(nm, "_16")]]
})
names(train_deltas) <- paste0(common_emb, "_delta")

train_shift <- bind_cols(train_join, train_deltas) %>%
  select(
    county_fips, county_name, state_po, vote_density,
    rep_share_16, rep_share_20,
    SHIFT_16_20,
    ends_with("_delta")
  ) %>%
  na.omit()

cat("\nTRAIN data loaded (2016->2020) for", nrow(train_shift), "counties.\n")
cat("Avg TRAIN shift:", round(mean(train_shift$SHIFT_16_20) * 100, 2), "points\n")

# ==========================================
# 2. TEST data: 2020 vs 2024 (delta = 2024 - 2020)
# ==========================================

df_20_base <- df_20 %>%
  select(
    county_fips,
    rep_share_20 = rep_share,
    all_of(common_emb)
  ) %>%
  rename_with(~ paste0(.x, "_20"), all_of(common_emb))

df_24_small <- df_24 %>%
  select(
    county_fips,
    county_name,
    state_po,
    vote_density,
    rep_share_24 = rep_share,
    all_of(common_emb)
  ) %>%
  rename_with(~ paste0(.x, "_24"), all_of(common_emb))

test_join <- inner_join(df_20_base, df_24_small, by = "county_fips") %>%
  mutate(SHIFT_20_24 = rep_share_24 - rep_share_20)

# Build delta predictors safely (NO cur_column)
test_deltas <- purrr::map_dfc(common_emb, function(nm) {
  test_join[[paste0(nm, "_24")]] - test_join[[paste0(nm, "_20")]]
})
names(test_deltas) <- paste0(common_emb, "_delta")

test_shift <- bind_cols(test_join, test_deltas) %>%
  select(
    county_fips, county_name, state_po, vote_density,
    rep_share_20,
    SHIFT_20_24,
    ends_with("_delta")
  ) %>%
  na.omit()

cat("\nTEST data loaded (2020->2024) for", nrow(test_shift), "counties.\n")
cat("Avg TEST shift:", round(mean(test_shift$SHIFT_20_24) * 100, 2), "points\n")

# ==========================================
# OPTIONAL: Remove extreme shift outliers (|SHIFT| > 0.15)
#   Drop this RIGHT AFTER you create train_shift and test_shift (after na.omit()).
#   Easy to comment out as a block.
# ==========================================

outlier_cutoff <- 0.15

# --- TRAIN outliers (2016 -> 2020) ---
n_train_before <- nrow(train_shift)
train_shift <- train_shift %>% filter(abs(SHIFT_16_20) <= outlier_cutoff)
n_train_after <- nrow(train_shift)

cat("\n[Outlier filter | TRAIN] Before:", n_train_before,
    "| After:", n_train_after,
    "| Removed:", (n_train_before - n_train_after), "\n")
cat("[Outlier filter | TRAIN] Shift range now:",
    round(min(train_shift$SHIFT_16_20), 4), "to",
    round(max(train_shift$SHIFT_16_20), 4), "\n")

# --- TEST outliers (2020 -> 2024) ---
n_test_before <- nrow(test_shift)
test_shift <- test_shift %>% filter(abs(SHIFT_20_24) <= outlier_cutoff)
n_test_after <- nrow(test_shift)

cat("\n[Outlier filter | TEST]  Before:", n_test_before,
    "| After:", n_test_after,
    "| Removed:", (n_test_before - n_test_after), "\n")
cat("[Outlier filter | TEST]  Shift range now:",
    round(min(test_shift$SHIFT_20_24), 4), "to",
    round(max(test_shift$SHIFT_20_24), 4), "\n")

# ==========================================
# 3. Ensure identical predictor columns between TRAIN and TEST
# ==========================================

id_cols_train <- c("county_fips", "county_name", "state_po", "vote_density",
                   "rep_share_16", "rep_share_20", "SHIFT_16_20")
id_cols_test  <- c("county_fips", "county_name", "state_po", "vote_density",
                   "rep_share_20", "SHIFT_20_24")

x_cols_train <- setdiff(names(train_shift), id_cols_train)
x_cols_test  <- setdiff(names(test_shift),  id_cols_test)

common_x <- intersect(x_cols_train, x_cols_test)

train_shift <- train_shift %>% select(all_of(id_cols_train), all_of(common_x))
test_shift  <- test_shift  %>% select(all_of(id_cols_test),  all_of(common_x))

cat("\nUsing", length(common_x), "shared DELTA predictors.\n")

# ==========================================
# 4. 5-Fold CV on TRAIN (2016->2020), predict TEST (2020->2024)
# ==========================================

set.seed(55)
train_shift <- train_shift %>%
  mutate(fold_id = sample(rep(1:5, length.out = n())))

test_preds_by_fold <- list()
importances <- list()

cat("\nStarting CV: train on 2016->2020 DELTAS, predict 2020->2024 DELTAS...\n")

for (k in 1:5) {

  train_df <- train_shift %>% filter(fold_id != k)

  rf_model <- ranger(
    SHIFT_16_20 ~ . - county_fips - county_name - state_po - vote_density - rep_share_16 - rep_share_20 - fold_id,
    data       = train_df,
    num.trees  = 500,
    mtry       = min(21, length(common_x)),  # guard if fewer predictors
    importance = "impurity",
    seed       = 42
  )

  preds <- predict(rf_model, data = test_shift)$predictions

  tmp <- test_shift %>%
    mutate(
      fold_model = k,
      Predicted_Shift = preds
    )

  test_preds_by_fold[[k]] <- tmp
  importances[[k]] <- rf_model$variable.importance

  cat("  Fold", k, "model trained (on 2016->2020) and applied to 2020->2024.\n")
}

# Average predictions across the 5 independently-trained models
results <- bind_rows(test_preds_by_fold) %>%
  group_by(county_fips, county_name, state_po, vote_density, rep_share_20, SHIFT_20_24) %>%
  summarise(
    Predicted_Shift = mean(Predicted_Shift),
    .groups = "drop"
  ) %>%
  mutate(
    Error = Predicted_Shift - SHIFT_20_24,
    Abs_Error = abs(Error)
  )

# ==========================================
# 5. Metrics (evaluated on 2020->2024)
# ==========================================

rss  <- sum(results$Error^2)
tss  <- sum((results$SHIFT_20_24 - mean(results$SHIFT_20_24))^2)
r2   <- 1 - (rss / tss)
rmse <- sqrt(mean(results$Error^2))

cat("\n==========================================")
cat("\nDELTA TRANSFER MODEL RESULTS")
cat("\nTrain: 2016->2020 (delta 20-16) | Test: 2020->2024 (delta 24-20)")
cat("\n==========================================")
cat("\nR-Squared: ", round(r2, 3))
cat("\nRMSE:      ", round(rmse, 4), "(", round(rmse*100, 2), "percentage points )")
cat("\n==========================================\n")

# ==========================================
# 6. Visualization 1: Actual vs Predicted Shift
# ==========================================

p1 <- ggplot(results, aes(x = SHIFT_20_24, y = Predicted_Shift)) +
  geom_point(alpha = 0.3, size = 1, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", color = "orange", se = FALSE) +
  labs(
    title = "Delta-embeddings transfer: (20-16) trained, (24-20) predicted",
    subtitle = paste0("R² = ", round(r2, 3), " | RMSE = ", round(rmse, 4)),
    x = "Actual Shift (2020 → 2024)",
    y = "Predicted Shift (trained on 2016 → 2020 deltas)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

print(p1)

# ==========================================
# 7. Visualization 2: Error vs Baseline Partisanship
# ==========================================

p2 <- ggplot(results, aes(x = rep_share_20, y = Error)) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_smooth(method = "loess", color = "blue", fill = "gray90") +
  labs(
    title = "Where were the surprises (delta transfer model)?",
    subtitle = "Positive Error = model expected a more pro-GOP shift than happened",
    x = "2020 GOP Vote Share (Baseline)",
    y = "Prediction Error (Pred - Actual)"
  ) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal()

print(p2)

# ==========================================
# 8. Visualization 3: Top Predictive Delta Features
# ==========================================

avg_imp <- bind_rows(lapply(importances, function(x) as.data.frame(t(x)))) %>%
  summarise(across(everything(), mean)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Importance") %>%
  arrange(desc(Importance)) %>%
  slice(1:15)

p3 <- ggplot(avg_imp, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Which delta embeddings predict swings (learned from 2016→2020)?",
    subtitle = "Top 15 delta features by impurity importance (avg across folds)",
    x = "",
    y = "Variable Importance (Impurity)"
  ) +
  theme_minimal()

print(p3)

# ==========================================
# 9. OPTIONAL DROP-IN VIS: R² by vote_density decile
#   - Put this AFTER you create `results` and compute r2/rmse (anywhere after results exists).
# ==========================================

results_r2_by_decile <- results %>%
  mutate(
    density_decile = ntile(vote_density, 10)
  ) %>%
  group_by(density_decile) %>%
  summarise(
    n = n(),
    rss = sum(Error^2),
    tss = sum((SHIFT_20_24 - mean(SHIFT_20_24))^2),
    r2  = 1 - rss / tss,
    rmse = sqrt(mean(Error^2)),
    density_min = min(vote_density, na.rm = TRUE),
    density_max = max(vote_density, na.rm = TRUE),
    .groups = "drop"
  )

p4 <- ggplot(results_r2_by_decile, aes(x = factor(density_decile), y = r2)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Model performance by vote-density decile (Delta transfer model)",
    subtitle = "R² computed within each decile on the 2020→2024 test set",
    x = "Vote-density decile (1 = lowest density, 10 = highest)",
    y = "R² within decile"
  ) +
  theme_minimal()

print(p4)

# ==========================================
# 10. Save
# ==========================================

write_csv(results, "data/results/county_change_m3_delta_results.csv")
cat("Saved: data/results/county_change_m3_delta_results.csv\n")

```
















# Let's validate the data!
```{r}

library(tidyverse)

df_16 <- read.csv("data/prep/counties/clean_county_emb_16_17.csv")
df_20 <- read.csv("data/prep/counties/clean_county_emb_20.csv")
df_24 <- read.csv("data/prep/counties/clean_county_emb_24.csv")

dups <- function(df, name) {
  n_dup <- df %>% count(county_fips) %>% filter(n > 1) %>% nrow()
  cat(name, "- duplicate county_fips rows:", n_dup, "\n")
}

dups(df_16, "df_16")
dups(df_20, "df_20")
dups(df_24, "df_24")

train_join <- inner_join(
  df_16 %>% select(county_fips, rep_share_16 = rep_share),
  df_20 %>% select(county_fips, rep_share_20 = rep_share),
  by = "county_fips"
)

test_join <- inner_join(
  df_20 %>% select(county_fips, rep_share_20 = rep_share),
  df_24 %>% select(county_fips, rep_share_24 = rep_share),
  by = "county_fips"
)

cat("Rows train_join:", nrow(train_join), "\n")
cat("Unique county_fips train_join:", n_distinct(train_join$county_fips), "\n\n")

cat("Rows test_join:", nrow(test_join), "\n")
cat("Unique county_fips test_join:", n_distinct(test_join$county_fips), "\n")

```
```{r}

summary(df_16$rep_share)
summary(df_20$rep_share)
summary(df_24$rep_share)

range(df_16$rep_share, na.rm = TRUE)
range(df_20$rep_share, na.rm = TRUE)
range(df_24$rep_share, na.rm = TRUE)

```

```{r}
train_shift_vals <- train_join %>% mutate(SHIFT_16_20 = rep_share_20 - rep_share_16)
test_shift_vals  <- test_join  %>% mutate(SHIFT_20_24 = rep_share_24 - rep_share_20)

summary(train_shift_vals$SHIFT_16_20)
summary(test_shift_vals$SHIFT_20_24)

sd(train_shift_vals$SHIFT_16_20, na.rm=TRUE)
sd(test_shift_vals$SHIFT_20_24, na.rm=TRUE)


shift <- test_shift_vals$SHIFT_20_24
quantile(shift, c(.001,.01,.05,.5,.95,.99,.999), na.rm=TRUE)
```
```{r}
library(tidyverse)

# --- Load 2020 + 2024 with totalvotes kept ---
df_20 <- read_csv("data/prep/counties/clean_county_emb_20.csv") %>%
  select(county_fips, county_name, state_po,
         rep_share_20 = rep_share,
         totalvotes_20 = total_votes)

df_24 <- read_csv("data/prep/counties/clean_county_emb_24.csv") %>%
  select(county_fips,
         rep_share_24 = rep_share,
         totalvotes_24 = total_votes)

# --- Join + compute shift ---
shift_tbl <- inner_join(df_20, df_24, by = "county_fips") %>%
  mutate(
    SHIFT_20_24 = rep_share_24 - rep_share_20,
    votes_min   = pmin(totalvotes_20, totalvotes_24),
    votes_max   = pmax(totalvotes_20, totalvotes_24)
  )

# --- Bottom 25 (biggest negative shifts) ---
bottom25 <- shift_tbl %>%
  arrange(SHIFT_20_24) %>%
  slice(1:25) %>%
  select(county_fips, county_name, state_po,
         rep_share_20, rep_share_24, SHIFT_20_24,
         totalvotes_20, totalvotes_24, votes_min, votes_max)

# --- Top 25 (biggest positive shifts) ---
top25 <- shift_tbl %>%
  arrange(desc(SHIFT_20_24)) %>%
  slice(1:25) %>%
  select(county_fips, county_name, state_po,
         rep_share_20, rep_share_24, SHIFT_20_24,
         totalvotes_20, totalvotes_24, votes_min, votes_max)

cat("\nBottom 25 shifts (most negative):\n")
print(bottom25)

cat("\nTop 25 shifts (most positive):\n")
print(top25)

```

```{r}

library(dplyr)

var = 20207
og <- read.csv("data/prep/counties/countypres_2000-2024.csv")

# Your original check (still fine)
og[og$county_fips == var, c("county_fips", "party", "year", "candidatevotes", "totalvotes", "mode")]


shift_tbl[shift_tbl$county_fips == var, ]
```

```{r}
library(dplyr)
library(stringr)

og <- read.csv("data/prep/counties/countypres_2000-2024.csv")

pres_2024 <- og %>%
  filter(year == 2024) %>%
  mutate(
    office_up = toupper(trimws(as.character(office))),
    candidate = str_squish(as.character(candidate)),
    party     = str_squish(as.character(party)),
    cand_up   = toupper(candidate),
    party_up  = toupper(party)
  ) %>%
  filter(grepl("PRESIDENT", office_up))

# define the same "summary row" flag
pres_2024 <- pres_2024 %>%
  mutate(
    is_summary_row =
      grepl("TOTAL", cand_up) |
      grepl("VOTES CAST", cand_up) |
      ((party_up == "" | is.na(party_up)) &
         !is.na(candidatevotes) & !is.na(totalvotes) &
         candidatevotes == totalvotes)
  )

# For each county_fips: do we have ANY non-summary candidate rows?
county_summary_status <- pres_2024 %>%
  group_by(county_fips) %>%
  summarise(
    n_rows = n(),
    n_non_summary = sum(!is_summary_row, na.rm = TRUE),
    .groups = "drop"
  )

# counties that would be fully dropped by summary-row removal
all_summary_counties <- county_summary_status %>%
  filter(n_non_summary == 0)

cat("2024 counties with ONLY summary rows:", nrow(all_summary_counties), "\n")
head(all_summary_counties, 20)
```

