---
title: "exploration"
format: html
---

```{r}

county <- read.csv("data/counties/2020_county_presidential.csv")
head(county)                          # checking the data loaded correctly
unique(county$state_name)             # just checking all 50 are present
unique(county$county_fips)    # checking number of unique counties

county[county$state_name == "Alaska", ]  # checking Alaska dat

a <- read.csv("data/counties/county_embeddings_2020.csv")
a
```

```{r}

## This code retrieves the shapefiles!


library(xml2)
library(rvest)

base_url <- "https://www2.census.gov/geo/tiger/TIGER2020PL/LAYER/COUNTY/2020/"

page <- read_html(base_url)
links <- html_elements(page, "a") |> html_attr("href")
zip_files <- links[grepl("\\.zip$", links)]

out_dir <- "data/counties/shapefiles"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# bump timeout
options(timeout = max(600, getOption("timeout")))

for (f in zip_files) {
  file_url  <- paste0(base_url, f)
  dest_file <- file.path(out_dir, f)

  # skip if already downloaded
  if (file.exists(dest_file)) {
    message("Skipping ", f, " (already exists)")
    next
  }

  message("Downloading ", f, " ...")

  tryCatch(
    {
      download.file(file_url, destfile = dest_file, mode = "wb", quiet = TRUE)
      message("  ✓ Finished ", f)
    },
    error = function(e) {
      message("  ✗ FAILED ", f, " : ", e$message)
    }
  )
}

```

```{r}

## This code unzips the shapefiles!
library(sf)
library(dplyr)
library(ggplot2)

zip_dir  <- "data/counties/shapefiles"
unzip_dir <- "data/counties/shapefiles_unzipped"

dir.create(unzip_dir, recursive = TRUE, showWarnings = FALSE)

# all the zip files we downloaded from Census
zip_files <- list.files(zip_dir, pattern = "\\.zip$", full.names = TRUE)

for (z in zip_files) {
  message("Unzipping ", basename(z), " ...")
  # put each state into its own subfolder (based on zip name)
  exdir_state <- file.path(
    unzip_dir,
    tools::file_path_sans_ext(basename(z))
  )
  dir.create(exdir_state, showWarnings = FALSE)
  unzip(z, exdir = exdir_state)
}

```

```{r}

# This code reads in all the shapefiles we just unzipped!
# find every county20 shapefile we just unzipped
shp_files <- list.files(
  unzip_dir,
  pattern = "county20\\.shp$",
  full.names = TRUE,
  recursive = TRUE
)

length(shp_files)   # should be ~56

alaska <- st_read("data/counties/shapefiles_unzipped/tl_2020_02_county20/tl_2020_02_county20.shp")
nrow(alaska)

# read & row-bind into one big sf
county_shapes <- do.call(
  rbind,
  lapply(shp_files, st_read, quiet = TRUE)
)

```
```{r}

# ---- 1. Prepare a lightweight shapefile ----
county_shapes_export <- county_shapes %>%
  mutate(county_fips = as.character(county_fips)) %>%
  select(county_fips, STATEFP20, geometry)

# Output folder
out_dir <- "data/counties"
shp_name <- "usa_counties"

# ---- 2. Write the shapefile (produces .shp, .dbf, .shx, .prj, .cpg) ----
st_write(
  county_shapes_export,
  dsn = file.path(out_dir, paste0(shp_name, ".shp")),
  delete_dsn = TRUE
)

# ---- 3. Zip all shapefile components together ----
# list all files with that prefix (e.g., usa_counties.*)
shp_files <- list.files(
  path = out_dir,
  pattern = paste0("^", shp_name, "\\.(shp|dbf|shx|prj|cpg)$"),
  full.names = TRUE
)

# Zip them into usa_counties.zip
zipfile_path <- file.path(out_dir, paste0(shp_name, ".zip"))
zip::zipr(zipfile_path, files = shp_files)



```


```{r}
library(dplyr)

# Alaska in the election data
county_ak <- county %>%
  filter(substr(county_fips, 1, 2) == "02") %>%
  count(county_fips, sort = TRUE)

nrow(county_ak)

# Alaska in the shapefile
shapes_ak <- county_shapes %>%
  filter(STATEFP20 == "02") %>%
  count(county_fips, sort = TRUE)

nrow(shapes_ak)

```


```{r}

library(stringr)

county <- county %>%
  mutate(
    county_fips = as.character(county_fips),
    # if some IDs are longer than 5 (e.g. county-splits), keep only the first 5
    county_fips = str_sub(county_fips, 1, 5),
    # then left-pad to 5 digits (in case anything came in shorter)
    county_fips = str_pad(county_fips, width = 5, side = "left", pad = "0")
  )


# shapefile: create same 5-digit FIPS from STATEFP20 + COUNTYFP20
county_shapes <- county_shapes %>%
  mutate(
    county_fips = paste0(STATEFP20, COUNTYFP20)
  )

# now attach ONLY the geometry column to the county data
county_sf <- county %>%
  left_join(
    county_shapes %>% select(STATEFP20, county_fips, geometry),
    by = "county_fips"
  ) %>%
  st_as_sf()

unique(substr(county$county_fips, 1, 2))            # should include "02"
unique(county_sf$STATEFP20)                         # should now also include "02"

```


```{r}
# ---- Packages ----
library(sf)
library(dplyr)
library(leaflet)
library(scales)

# county_sf: your joined sf object with per_dem in [0,1]

# ---- 1. Clean geometry and transform to WGS84 ----
county_poly <- county_sf %>%
  st_make_valid() %>%                       # fix invalid geometries if any
  st_cast("MULTIPOLYGON", warn = FALSE) %>% # ensure polygonal geometry
  st_transform(4326)                        # Leaflet wants EPSG:4326 (lat/long)

# ---- 2. Clamp per_dem so 15%/85% are full red/blue ----
county_poly <- county_poly %>%
  mutate(
    per_dem_clamped = pmin(pmax(per_dem, 0.15), 0.85)
  )

# ---- 3. Build a robust palette domain ----
vals <- county_poly$per_dem_clamped
vals_finite <- vals[is.finite(vals) & !is.na(vals)]

pal <- colorNumeric(
  palette  = c("red", "white", "blue"),
  domain   = range(vals_finite, na.rm = TRUE),
  na.color = "transparent"
)

# ---- 4. Interactive Leaflet map (no labels yet) ----
leaflet(county_poly, options = leafletOptions(minZoom = 3)) %>%
  addTiles() %>%  # OpenStreetMap basemap
  addPolygons(
    fillColor   = ~pal(per_dem_clamped),
    fillOpacity = 0.8,
    color       = "#555555",   # county borders
    weight      = 0.1,
    opacity     = 0.3,
    smoothFactor = 0.1,
    highlightOptions = highlightOptions(
      weight = 1,
      color = "black",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal      = pal,
    values   = vals_finite,
    title    = "Democratic share",
    opacity  = 0.8
  )

```


```{r}

ggplot(county_sf) +
  geom_sf(aes(fill = per_dem), color = NA) +
  scale_fill_gradient2(
    low = "red",       # Republican counties
    mid = "white",     # Even split (≈50%)
    high = "blue",     # Democratic counties
    midpoint = 0.5,    # Center the scale at 50% Dem share
    limits = c(0.15, 0.85),
    oob = scales::squish, 
    na.value = "grey80",
    name = "Democratic share"
  ) +
  theme_void() +
  labs(
    title = "US Counties – 2020 Presidential Vote Share",
    subtitle = "Blue = more Democratic, Red = more Republican"
  )




```



```{r}


library(sf)        # for shapefiles
library(dplyr)     # for joins / mutate
library(ggplot2)   # for plotting

shp_path <- "data/tl_2020_01_county20/tl_2020_01_county20.shp"

shp <- st_read(shp_path)

# county: make sure county_fips is a zero-padded 5-character string
county <- county %>%
  mutate(county_fips = sprintf("%05d", county_fips))

# shapefile: build 5-digit county_fips from STATEFP + COUNTYFP
# (if you already have a 5-digit GEOID, you can reuse that instead)
shp <- shp %>%
  mutate(county_fips = paste0(STATEFP20, COUNTYFP20))

county_sf <- county %>%
  left_join(
    shp %>% select(county_fips, geometry),
    by = "county_fips"
  ) %>%
  st_as_sf()   # make sure the result remains an sf object

ggplot(county_sf) +
  geom_sf(aes(fill = per_dem), color = NA) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey80") +
  theme_void() +
  labs(fill = "Dem share",
       title = "Test map: election data joined to county shapefile")


```

```{r}

nyt <- st_read("data/precincts/nyt_2024_shapefiles_primarykey/precincts_id.shp")
head(nyt)
nrow(nyt)

ak <- st_read("data/mixed/unzipped_pkey/ak/ak_2020.shp")

test <- read.csv("data/prep/precincts/processed_area_std_uuid.csv")


```









```{r}

# load libraries
library(dplyr)
library(readr)
library(ranger)
library(tidymodels)
library(yardstick)
library(xgboost)
library(tidyverse)
library(sf)
library(uuid)

```

Process the original mean precinct data
```{r}

mixed <- read.csv("data/mixed/mixed_embeddings_2020.csv")

mixed <- mixed %>%
  mutate(TRUMPSHARE = G20PRERTRU / TOTALVOTES) %>%
  select(state, TOTALVOTES, TRUMPSHARE, G20PREDBID, G20PRERTRU, system.index, starts_with("A"))

write_csv(mixed, "data/mixed/processed_mixed_embeddings_2020.csv")

```

Temporary because no pkey -- just to get an initial look at predictive power
```{r}

mixed <- read.csv("data/mixed/precinctarea_embeddings_2020.csv")
head(mixed)

mixed <- mixed %>%
  mutate(TRUMPSHARE = G20PRERTRU / TOTALVOTES) %>%
  select(state, precinct_area, TOTALVOTES, TRUMPSHARE, G20PREDBID, G20PRERTRU, system.index, starts_with("A"))

write_csv(mixed, "data/mixed/processed_mixedarea_embeddings_2020.csv")

```


Process mean precinct data 2020 with primary key
```{r}

mixed_pkey <- read.csv("data/mixed/precinct_embeddings_2020_pkey.csv")

mixed_pkey <- mixed_pkey[mixed_pkey$TOTALVOTES >= 5, ]   # filter out tiny precincts

mixed_pkey <- mixed_pkey %>%
  mutate(TRUMPSHARE = G20PRERTRU / TOTALVOTES) %>%
  mutate(BIDENSHARE = G20PREDBID / TOTALVOTES) %>%
  select(UUIDPKEY, state, TOTALVOTES, TRUMPSHARE, BIDENSHARE, G20PREDBID, G20PRERTRU, system.index, starts_with("A"))

write_csv(mixed_pkey, "data/mixed/2020_precinct_embeddings_uuid.csv")

```


PROCESS 2024 NYT EMBEDDINGS
```{r}

df <- read.csv("data/precincts/nyt_2024_embeddings.csv")
head(df)
nrow(df)

sum(df$votes_tota)    # number of votes is 153,850,356 -- looks right!

nrow(df[df$votes_tota == 0, ])    # 0 precincts with no votes
nrow(df[df$votes_tota < 10, ])    # 1577 precincts with fewer than ten voters

sum(df[df$votes_tota < 10, ]$votes_tota) # 6324 votes in precincts with fewer than 10 total votes

df1 <- df %>%
  mutate(REPUBSHARE = votes_rep / votes_tota) %>%
  mutate(DEMOCSHARE = votes_dem / votes_tota) %>%
  mutate(TOTALVOTES = votes_tota) %>%
  mutate(REPUBVOTES = votes_rep) %>%
  mutate(DEMOCVOTES = votes_dem) %>%
  select(UUIDPKEY, state, TOTALVOTES, REPUBSHARE, DEMOCSHARE, DEMOCVOTES, REPUBVOTES, system.index, starts_with("A"))

df1 <- df1[df1$TOTALVOTES >= 10, ]   # filter out tiny precincts

write_csv(df1, "data/precincts/processed_2024_embeddings.csv")

```


Process the stdev precinct data 2020
```{r}
mixed_sd <- read.csv("data/precincts/precinct_embeddings_2020_mean_std_skew.csv")
head(mixed_sd)

mixed_sd <- mixed_sd %>%
  mutate(TRUMPSHARE = G20PRERTRU / TOTALVOTES) %>%
  select(state, TOTALVOTES, TRUMPSHARE, G20PREDBID, G20PRERTRU, system.index, starts_with("A"))

write_csv(mixed_sd, "data/mixed/sd_processed_mixed_embeddings_2020_sd.csv")
```

PROCESS STD + SKEW DATA 2020 KEEP UUID AND PRECINCT AREA
```{r}
mixed_std <- read.csv("data/mixed/precinct_embeddings_2020_mean_area_std.csv")
head(mixed_std)

mixed_std <- mixed_std %>%
  mutate(TRUMPSHARE = G20PRERTRU / TOTALVOTES) %>%
  select(state, TOTALVOTES, TRUMPSHARE, G20PREDBID, G20PRERTRU, precinct_area, UUIDPKEY, starts_with("A"))

write_csv(mixed_std, "data/mixed/processed_area_std_uuid.csv")
```

```{r}
csv <- read.csv("data/mixed/processed_area_std_uuid.csv")
colnames(csv)
head(csv)
```



READ IN PRECINCT DATA!
```{r}

df <- read.csv("data/mixed/processed_mixed_embeddings_2020.csv")
dfsd <- read.csv("data/mixed/sd_processed_mixed_embeddings_2020_sd.csv")
dfarea <- read.csv("data/mixed/processed_mixedarea_embeddings_2020.csv")

```

Filter out tiny precincts!
```{r}

head(df[,c("TOTALVOTES","TRUMPSHARE","G20PREDBID","G20PRERTRU")])

sum(df$TOTALVOTES)    # number of votes is 158,425,815 -- looks right!

nrow(df[df$TOTALVOTES == 0, ])    # 6,296 precincts with no votes
sum(df[df$TOTALVOTES < 5, ]$TOTALVOTES)   # only 3,510 votes total from precincts with fewer than 5 votes

df2 <- df[df$TOTALVOTES >= 5, ]   # filter out tiny precincts
dfsd2 <- dfsd[dfsd$TOTALVOTES >= 5, ]   # filter out tiny precincts for the sd set
dfarea2 <- dfarea[dfarea$TOTALVOTES >= 5, ]   # filter out tiny precincts for the area set
```

```{r}

# ==========================================
# Random forest with ranger on embeddings
# ==========================================

set.seed(10)  # for reproducibility

# Keep only the columns we need and drop NAs
#    (assumes A00:A63 exist and are the 64 embedding dims)
data_rf <- csv %>%
  select(TRUMPSHARE, precinct_area, TOTALVOTES) %>%
  na.omit()

# Train/test split (80/20)
n <- nrow(data_rf)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- data_rf[train_idx, ]
test_df  <- data_rf[-train_idx, ]

cat("Train rows:", nrow(train_df), " | Test rows:", nrow(test_df), "\n")

# Fit ranger random forest
rf_model <- ranger(
  TRUMPSHARE ~ .,
  data       = train_df,
  num.trees  = 500,
  mtry       = 2,
  importance = "impurity",   # or "permutation" if you want stricter importance
  seed       = 42
)

# Print a brief model summary (includes OOB R²)
print(rf_model)

# Predict on the test set
rf_pred <- predict(rf_model, data = test_df)$predictions

# Evaluate performance on the test set
rss  <- sum((rf_pred - test_df$TRUMPSHARE)^2)
tss  <- sum((test_df$TRUMPSHARE - mean(test_df$TRUMPSHARE))^2)
r2   <- 1 - rss / tss
rmse <- sqrt(mean((rf_pred - test_df$TRUMPSHARE)^2))

cat("\nPerformance on test set:\n")
cat("R²   =", round(r2, 3), "\n")
cat("RMSE =", round(rmse, 4), "\n\n")

# Variable importance summary
imp <- sort(rf_model$variable.importance, decreasing = TRUE)

cat("Top 10 most important features:\n")
print(round(imp[1:10], 4))


```


LINEAR REGRESSION
```{r}

formula <- as.formula(
  paste("TRUMPSHARE ~", paste(paste0("A", sprintf("%02d", 1:63)), collapse = " + "))
)

model <- lm(formula, data = df2)
summary(model)

```

```{r}

summary(lm(TRUMPSHARE ~ TOTALVOTES + precinct_area + state, data = csv))

```



Random Forest with default 500 trees and minimum node size of 5
```{r}
# ==========================================
# Random forest with ranger on embeddings
# ==========================================

set.seed(10)  # for reproducibility

# Keep only the columns we need and drop NAs
#    (assumes A00:A63 exist and are the 64 embedding dims)
data_rf <- df2 %>%
  select(TRUMPSHARE, A00:A63) %>%
  na.omit()

# Train/test split (80/20)
n <- nrow(data_rf)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- data_rf[train_idx, ]
test_df  <- data_rf[-train_idx, ]

cat("Train rows:", nrow(train_df), " | Test rows:", nrow(test_df), "\n")

# Fit ranger random forest
rf_model <- ranger(
  TRUMPSHARE ~ .,
  data       = train_df,
  num.trees  = 500,
  mtry       = 8,
  importance = "impurity",   # or "permutation" if you want stricter importance
  seed       = 42
)

# Print a brief model summary (includes OOB R²)
print(rf_model)

# Predict on the test set
rf_pred <- predict(rf_model, data = test_df)$predictions

# Evaluate performance on the test set
rss  <- sum((rf_pred - test_df$TRUMPSHARE)^2)
tss  <- sum((test_df$TRUMPSHARE - mean(test_df$TRUMPSHARE))^2)
r2   <- 1 - rss / tss
rmse <- sqrt(mean((rf_pred - test_df$TRUMPSHARE)^2))

cat("\nPerformance on test set:\n")
cat("R²   =", round(r2, 3), "\n")
cat("RMSE =", round(rmse, 4), "\n\n")

# Variable importance summary
imp <- sort(rf_model$variable.importance, decreasing = TRUE)

cat("Top 10 most important features:\n")
print(round(imp[1:10], 4))

```
Test set performance (set seed 42): 
R squared: 0.736
RMSE: 0.1203

Test set performance (set seed 43): 
R squared: 0.733
RMSE: 0.121

Test set performance (set seed 10): 
R squared: 0.734
RMSE: 0.1206

WITH THE NEW DATA (newer version, same data) let's see what we get
```{r}

library(tidyverse)
library(ranger)

# ==========================================
# 1. Load and Prepare Data
# ==========================================

# Load the new clean 2020 dataset
clean_data <- read_csv("data/prep/precincts/2020/clean_emb_precinct_20.csv")

# Prepare data for RF:
# 1. Rename rep_share -> TRUMPSHARE to match your existing RF code
# 2. Select TRUMPSHARE and ONLY the columns ending in "_mean" (the embedding means)
data_rf <- clean_data %>%
  select(TRUMPSHARE = rep_share, ends_with("_mean")) %>%
  na.omit()

# Check to ensure we have the right dimensions (Expect ~65 columns: 1 target + 64 features)
cat("Data dimensions:", dim(data_rf)[1], "rows x", dim(data_rf)[2], "columns\n")

# ==========================================
# 2. Random forest with ranger (UNMODIFIED)
# ==========================================

set.seed(10)  # for reproducibility

# Train/test split (80/20)
n <- nrow(data_rf)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- data_rf[train_idx, ]
test_df  <- data_rf[-train_idx, ]

cat("Train rows:", nrow(train_df), " | Test rows:", nrow(test_df), "\n")

# Fit ranger random forest
rf_model <- ranger(
  TRUMPSHARE ~ .,
  data       = train_df,
  num.trees  = 500,
  mtry       = 8,
  importance = "impurity",   # or "permutation" if you want stricter importance
  seed       = 42
)

# Print a brief model summary (includes OOB R²)
print(rf_model)

# Predict on the test set
rf_pred <- predict(rf_model, data = test_df)$predictions

# Evaluate performance on the test set
rss  <- sum((rf_pred - test_df$TRUMPSHARE)^2)
tss  <- sum((test_df$TRUMPSHARE - mean(test_df$TRUMPSHARE))^2)
r2   <- 1 - rss / tss
rmse <- sqrt(mean((rf_pred - test_df$TRUMPSHARE)^2))

cat("\nPerformance on test set:\n")
cat("R²   =", round(r2, 3), "\n")
cat("RMSE =", round(rmse, 4), "\n\n")

# Variable importance summary
imp <- sort(rf_model$variable.importance, decreasing = TRUE)

cat("Top 10 most important features:\n")
print(round(imp[1:10], 4))

```


Random forest on larger dataset with mean, stdDev, and skew features
```{r}

# =========================================================================
# Random forest on Extended Embeddings (Mean, StdDev, Skew)
# =========================================================================

set.seed(42)  # for reproducibility

# 1. Prepare Data
# -------------------------------------------------------------------------
# Select TRUMPSHARE and all columns that end in mean, stdDev, or skew.
# This avoids hardcoding ranges like A00:A63
data_rf <- dfsd2 %>%
  select(TRUMPSHARE, matches("mean|stdDev|skew")) %>%
  na.omit()

# Check dimensions to ensure we have approx 192 predictors + 1 outcome
cat("Data dimensions:", dim(data_rf)[1], "rows and", dim(data_rf)[2], "columns\n")

# 2. Train/Test Split (80/20)
# -------------------------------------------------------------------------
n <- nrow(data_rf)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- data_rf[train_idx, ]
test_df  <- data_rf[-train_idx, ]

cat("Train rows:", nrow(train_df), " | Test rows:", nrow(test_df), "\n")

# 3. Fit Ranger Random Forest
# -------------------------------------------------------------------------
# We calculate a dynamic mtry. Standard for regression is p/3.
# With ~192 vars, mtry should be around 64.
num_features <- ncol(train_df) - 1
optimal_mtry <- floor(num_features / 3)

rf_model <- ranger(
  TRUMPSHARE ~ .,
  data       = train_df,
  num.trees  = 500,
  mtry       = optimal_mtry, 
  importance = "impurity",
  seed       = 42,
  num.threads = parallel::detectCores() - 1 # Use all cores but one for speed
)

# Print model summary
print(rf_model)

# 4. Evaluate Performance
# -------------------------------------------------------------------------
rf_pred <- predict(rf_model, data = test_df)$predictions

rss  <- sum((rf_pred - test_df$TRUMPSHARE)^2)
tss  <- sum((test_df$TRUMPSHARE - mean(test_df$TRUMPSHARE))^2)
r2   <- 1 - rss / tss
rmse <- sqrt(mean((rf_pred - test_df$TRUMPSHARE)^2))

cat("\n==============================\n")
cat("Performance on test set:\n")
cat("==============================\n")
cat("R²   =", round(r2, 3), "\n")
cat("RMSE =", round(rmse, 4), "\n")

# 5. Variable Importance
# -------------------------------------------------------------------------
imp <- sort(rf_model$variable.importance, decreasing = TRUE)

cat("\nTop 15 most important features:\n")
print(round(imp[1:15], 4))

# Optional: Check if StdDev or Skew provide value compared to Mean
# This calculates the % of the total importance held by each type of stat
mean_imp <- sum(imp[grep("mean", names(imp))])
std_imp  <- sum(imp[grep("stdDev", names(imp))])
skew_imp <- sum(imp[grep("skew", names(imp))])
total_imp <- sum(imp)

cat("\nImportance contribution by statistic type:\n")
cat("Mean:   ", round((mean_imp/total_imp)*100, 1), "%\n")
cat("StdDev: ", round((std_imp/total_imp)*100, 1), "%\n")
cat("Skew:   ", round((skew_imp/total_imp)*100, 1), "%\n")
```


Random Forest with 1000 trees and minimum node size of 3
```{r}

set.seed(42)  # for reproducibility

# Keep only the columns we need and drop NAs
#    (assumes A00:A63 exist and are the 64 embedding dims)
data_rf <- df2 %>%
  select(TRUMPSHARE, A00:A63) %>%
  na.omit()

# Train/test split (80/20)
n <- nrow(data_rf)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- data_rf[train_idx, ]
test_df  <- data_rf[-train_idx, ]

cat("Train rows:", nrow(train_df), " | Test rows:", nrow(test_df), "\n")

# Fit ranger random forest
rf_model <- ranger(
  TRUMPSHARE ~ .,
  data       = train_df,
  num.trees  = 1000,
  mtry       = 8,
  min.node.size = 3,
  importance = "impurity",   # or "permutation" if you want stricter importance
  seed       = 42
)

# Print a brief model summary (includes OOB R²)
print(rf_model)

# Predict on the test set
rf_pred <- predict(rf_model, data = test_df)$predictions

# Evaluate performance on the test set
rss  <- sum((rf_pred - test_df$TRUMPSHARE)^2)
tss  <- sum((test_df$TRUMPSHARE - mean(test_df$TRUMPSHARE))^2)
r2   <- 1 - rss / tss
rmse <- sqrt(mean((rf_pred - test_df$TRUMPSHARE)^2))

cat("\nPerformance on test set:\n")
cat("R²   =", round(r2, 3), "\n")
cat("RMSE =", round(rmse, 4), "\n\n")

# Variable importance summary
imp <- sort(rf_model$variable.importance, decreasing = TRUE)

cat("Top 10 most important features:\n")
print(round(imp[1:10], 4))

```
Test set performance: 
R squared: 0.737
RMSE: 0.1201


First pass XGBoost model -- doesn't perform as well
```{r}
# ==========================================
# XGBoost on embeddings for TRUMPSHARE
# ==========================================

set.seed(42)  # reproducibility

# 1️⃣ Ensure embeddings are numeric and select columns
df2_clean <- df2 %>%
  mutate(across(starts_with("A"), ~ as.numeric(.x)))

data_xgb <- df2_clean %>%
  select(TRUMPSHARE, A00:A63) %>%  # adjust if your cols differ
  na.omit()

cat("Rows after na.omit:", nrow(data_xgb), "\n")

# 2️⃣ Train/test split (80/20)
n <- nrow(data_xgb)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- data_xgb[train_idx, ]
test_df  <- data_xgb[-train_idx, ]

y_train <- train_df$TRUMPSHARE
y_test  <- test_df$TRUMPSHARE

X_train <- as.matrix(train_df[, -1])  # drop target column
X_test  <- as.matrix(test_df[, -1])

# 3️⃣ Create DMatrix objects (xgboost's optimized data structure)
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

watchlist <- list(train = dtrain, eval = dtest)

# 4️⃣ Set XGBoost parameters
params <- list(
  booster          = "gbtree",
  objective        = "reg:squarederror",  # regression with squared error loss
  eta              = 0.05,                # learning rate (shrinkage)
  max_depth        = 6,                   # depth of each tree
  subsample        = 0.8,                 # row subsampling
  colsample_bytree = 0.8,                 # feature subsampling
  min_child_weight = 1,                   # min sum of instance weight in a leaf
  lambda           = 1.0,                 # L2 regularization
  alpha            = 0.0,                 # L1 regularization
  eval_metric      = "rmse"               # track RMSE on eval set
)

# 5️⃣ Train the model with early stopping
set.seed(42)
xgb_model <- xgb.train(
  params                = params,
  data                  = dtrain,
  nrounds               = 500,     # max number of boosting rounds
  watchlist             = watchlist,
  early_stopping_rounds = 50,      # stop if no improvement for 50 rounds
  print_every_n         = 25       # how often to print progress
)

cat("\nBest iteration:", xgb_model$best_iteration, "\n")

# 6️⃣ Predict on the test set
xgb_pred <- predict(xgb_model, newdata = X_test)

# 7️⃣ Evaluate performance
rss  <- sum((xgb_pred - y_test)^2)
tss  <- sum((y_test - mean(y_test))^2)
r2   <- 1 - rss / tss
rmse <- sqrt(mean((xgb_pred - y_test)^2))

cat("\nXGBoost performance on test set:\n")
cat("R²   =", round(r2, 3), "\n")
cat("RMSE =", round(rmse, 4), "\n\n")

# 8️⃣ Variable importance
imp <- xgb.importance(
  feature_names = colnames(X_train),
  model         = xgb_model
)

cat("Top 10 most important features:\n")
print(imp[1:10, ])

```
Test set performance: 
R squared: 0.705
RMSE: 0.1271


Using tidymodels to optimize the parameter tuning -- reduce training sample to increase speed, tree depth, etc.
```{r}

set.seed(42)

# library(doParallel)  # optional if you want parallelization

# ==========================================================
# 1. Clean data: ensure embeddings are numeric & select cols
# ==========================================================

df2_clean <- df2 %>%
  mutate(across(starts_with("A"), ~ as.numeric(.x)))

data_xgb <- df2_clean %>%
  select(TRUMPSHARE, A00:A63) %>%
  na.omit()

cat("Rows after na.omit:", nrow(data_xgb), "\n")

# ==========================================================
# 2. Train/test split (80/20)
# ==========================================================

data_split <- initial_split(data_xgb, prop = 0.8, strata = TRUMPSHARE)
train_df   <- training(data_split)
test_df    <- testing(data_split)

# ==========================================================
# 3. Smaller tuning sample from training set
#    (Fit final model later on full train_df)
# ==========================================================

max_tune_n <- 50000L  # adjust up/down depending on patience
tune_df <- if (nrow(train_df) > max_tune_n) {
  train_df %>% slice_sample(n = max_tune_n)
} else {
  train_df
}

cat("Rows used for tuning:", nrow(tune_df), "\n")

# ==========================================================
# 4. Resampling: 3-fold CV on the tuning subset
# ==========================================================

set.seed(42)
folds <- vfold_cv(tune_df, v = 3, strata = TRUMPSHARE)

# ==========================================================
# 5. Recipes (one for tuning, one for final fit)
# ==========================================================

xgb_recipe_tune  <- recipe(TRUMPSHARE ~ ., data = tune_df)
xgb_recipe_full  <- recipe(TRUMPSHARE ~ ., data = train_df)

# ==========================================================
# 6. Model spec: XGBoost with tunable hyperparameters
#    (slightly narrower ranges for speed)
# ==========================================================

xgb_spec <- boost_tree(
  trees         = tune(),     # nrounds
  tree_depth    = tune(),     # max_depth
  learn_rate    = tune(),     # eta
  loss_reduction = tune(),    # gamma
  min_n         = tune(),     # min child size
  mtry          = tune(),     # colsample_bytree-ish
  sample_size   = 0.8         # fixed subsample
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Workflows for tuning & final fit
xgb_wf_tune <- workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(xgb_recipe_tune)

xgb_wf_full <- workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(xgb_recipe_full)

# ==========================================================
# 7. Define tuning parameter ranges (tighter + smaller grid)
# ==========================================================

predictor_df <- tune_df %>% select(-TRUMPSHARE)

xgb_params <- parameters(
  finalize(mtry(), predictor_df),            # 1..p
  trees(range = c(300L, 800L)),              # fewer trees than before
  tree_depth(range = c(3L, 6L)),             # shallower trees
  learn_rate(range = c(0.05, 0.2)),          # reasonable learning rate
  loss_reduction(),                          # default gamma range
  min_n(range = c(5L, 40L))                  # min child size
)

set.seed(42)
xgb_grid <- grid_space_filling(
  xgb_params,
  size = 10                                   # 10 configs instead of 25
)

# ==========================================================
# 8. Tune: CV performance for each grid point
# ==========================================================

metrics_to_use <- metric_set(yardstick::rmse, yardstick::rsq)

# Optional parallelization (uncomment if you want it and have doParallel)
# cl <- parallel::makeCluster(parallel::detectCores() - 1)
# doParallel::registerDoParallel(cl)

set.seed(42)
xgb_tune_res <- tune_grid(
  xgb_wf_tune,
  resamples = folds,
  grid = xgb_grid,
  metrics = metrics_to_use,
  control = control_grid(
    save_pred = FALSE,
    verbose = TRUE
  )
)

# If you used parallel:
# parallel::stopCluster(cl)

show_best(xgb_tune_res, metric = "rmse", n = 5)

# ==========================================================
# 9. Select best hyperparameters and finalize workflow
# ==========================================================

best_xgb <- select_best(xgb_tune_res, metric = "rmse")
best_xgb

final_xgb_wf <- finalize_workflow(xgb_wf_full, best_xgb)

# ==========================================================
# 🔟 Fit final model on full training data, evaluate on test
# ==========================================================

final_xgb_fit <- last_fit(final_xgb_wf, split = data_split)

# Performance on held-out test set
collect_metrics(final_xgb_fit)

# Predictions if you want them
xgb_test_preds <- collect_predictions(final_xgb_fit)
head(xgb_test_preds)

# Inspect the fitted xgboost model
final_xgb_model <- extract_fit_parsnip(final_xgb_fit$.workflow[[1]])
final_xgb_model

collect_metrics(final_xgb_fit)

```

Significantly worse performance:
RMSE: 0.15
R squared: 0.592

Second attempt using tidymodels -- more juice
```{r}

library(tidymodels)
# library(doParallel)  # optional if you want parallel

set.seed(42)

# ==========================================================
# 1. Clean data: embeddings numeric & select cols
# ==========================================================

df2_clean <- df2 %>%
  mutate(across(starts_with("A"), ~ as.numeric(.x)))

data_xgb <- df2_clean %>%
  select(TRUMPSHARE, A00:A63) %>%
  na.omit()

cat("Rows after na.omit:", nrow(data_xgb), "\n")

# ==========================================================
# 2. Train/test split (80/20)
# ==========================================================

data_split <- initial_split(data_xgb, prop = 0.8, strata = TRUMPSHARE)
train_df   <- training(data_split)
test_df    <- testing(data_split)

# ==========================================================
# 3. Tuning subset from training set
#    (XGBoost tuned on subset, final model on full train_df)
# ==========================================================

max_tune_n <- 120000L   # increase if you want, lower if too slow

tune_df <- if (nrow(train_df) > max_tune_n) {
  train_df %>% slice_sample(n = max_tune_n)
} else {
  train_df
}

cat("Rows used for tuning:", nrow(tune_df), "\n")

# ==========================================================
# 4. Resampling: 3-fold CV on tuning subset
# ==========================================================

set.seed(42)
folds <- vfold_cv(tune_df, v = 3, strata = TRUMPSHARE)

# ==========================================================
# 5. Recipes (tuning + final)
# ==========================================================

xgb_recipe_tune <- recipe(TRUMPSHARE ~ ., data = tune_df)
xgb_recipe_full <- recipe(TRUMPSHARE ~ ., data = train_df)

# ==========================================================
# 6. XGBoost model spec with more expressive trees
# ==========================================================

xgb_spec <- boost_tree(
  trees          = tune(),  # nrounds
  tree_depth     = tune(),  # max_depth
  learn_rate     = tune(),  # eta
  loss_reduction = tune(),  # gamma
  min_n          = tune(),  # min_child_weight-ish
  mtry           = tune(),  # colsample_bytree-ish
  sample_size    = 0.8      # subsample
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_wf_tune <- workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(xgb_recipe_tune)

xgb_wf_full <- workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(xgb_recipe_full)

# ==========================================================
# 7. Parameter ranges (deeper trees + more trees)
# ==========================================================

predictor_df <- tune_df %>% select(-TRUMPSHARE)

xgb_params <- parameters(
  finalize(mtry(), predictor_df),             # 1..p
  trees(range = c(800L, 2000L)),              # more trees
  tree_depth(range = c(4L, 12L)),             # deeper trees
  learn_rate(range = c(0.03, 0.15)),          # conservative learning rate
  loss_reduction(),                           # gamma (default range)
  min_n(range = c(5L, 50L))                   # min child size
)

set.seed(42)
xgb_grid <- grid_space_filling(
  xgb_params,
  size = 15                                   # can bump to 20 if you want
)

# ==========================================================
# 8. Tune with CV
# ==========================================================

metrics_to_use <- metric_set(yardstick::rmse, yardstick::rsq)

# Optional parallelization
# cl <- parallel::makeCluster(parallel::detectCores() - 1)
# doParallel::registerDoParallel(cl)

set.seed(42)
xgb_tune_res <- tune_grid(
  xgb_wf_tune,
  resamples = folds,
  grid = xgb_grid,
  metrics = metrics_to_use,
  control = control_grid(
    save_pred = FALSE,
    verbose = TRUE
  )
)

# if using parallel:
# parallel::stopCluster(cl)

cat("Best configs (by RMSE):\n")
show_best(xgb_tune_res, metric = "rmse", n = 5)

# ==========================================================
# 9. Select best hyperparameters & finalize workflow
# ==========================================================

best_xgb <- select_best(xgb_tune_res, metric = "rmse")
best_xgb

final_xgb_wf <- finalize_workflow(xgb_wf_full, best_xgb)

# ==========================================================
# 🔟 Fit final model on full training data, eval on test
# ==========================================================

final_xgb_fit <- last_fit(final_xgb_wf, split = data_split)

# Test-set performance (RMSE + R^2)
final_metrics <- collect_metrics(final_xgb_fit)
final_metrics

cat("\nTest-set R^2:\n")
final_metrics %>%
  filter(.metric == "rsq") %>%
  print()

# ==========================================================
# 11. Predictions & fitted model object (optional)
# ==========================================================

xgb_test_preds <- collect_predictions(final_xgb_fit)
head(xgb_test_preds)

final_xgb_model <- extract_fit_parsnip(final_xgb_fit$.workflow[[1]])
final_xgb_model


```







VERY BOTTOM, JUST A TEST FOR AREA -- FIRST, WITHOUT AREA, USING 150m resolution
WITHOUT FILTERING OUT THE TINY PRECINCTS, we get 0.711 performance on test set 
```{r}
# ==========================================
# Random forest with ranger on embeddings
# ==========================================

set.seed(10)  # for reproducibility

# Keep only the columns we need and drop NAs
#    (assumes A00:A63 exist and are the 64 embedding dims)
data_rf <- dfarea2 %>%
  select(TRUMPSHARE, A00:A63) %>%
  na.omit()

# Train/test split (80/20)
n <- nrow(data_rf)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- data_rf[train_idx, ]
test_df  <- data_rf[-train_idx, ]

cat("Train rows:", nrow(train_df), " | Test rows:", nrow(test_df), "\n")

# Fit ranger random forest
rf_model <- ranger(
  TRUMPSHARE ~ .,
  data       = train_df,
  num.trees  = 500,
  mtry       = 8,
  importance = "impurity",   # or "permutation" if you want stricter importance
  seed       = 42
)

# Print a brief model summary (includes OOB R²)
print(rf_model)

# Predict on the test set
rf_pred <- predict(rf_model, data = test_df)$predictions

# Evaluate performance on the test set
rss  <- sum((rf_pred - test_df$TRUMPSHARE)^2)
tss  <- sum((test_df$TRUMPSHARE - mean(test_df$TRUMPSHARE))^2)
r2   <- 1 - rss / tss
rmse <- sqrt(mean((rf_pred - test_df$TRUMPSHARE)^2))

cat("\nPerformance on test set:\n")
cat("R²   =", round(r2, 3), "\n")
cat("RMSE =", round(rmse, 4), "\n\n")

# Variable importance summary
imp <- sort(rf_model$variable.importance, decreasing = TRUE)

cat("Top 10 most important features:\n")
print(round(imp[1:10], 4))

```
0.736 R squared test set!

NEXT, WITH AREA, JUST TO SEE THE DIFFERENCE!
```{r}
# ==========================================
# Random forest with ranger on embeddings
# ==========================================

set.seed(10)  # for reproducibility

# Keep only the columns we need and drop NAs
#    (assumes A00:A63 exist and are the 64 embedding dims)
data_rf <- dfarea2 %>%
  select(TRUMPSHARE, precinct_area, A00:A63) %>%
  na.omit()

# Train/test split (80/20)
n <- nrow(data_rf)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))

train_df <- data_rf[train_idx, ]
test_df  <- data_rf[-train_idx, ]

cat("Train rows:", nrow(train_df), " | Test rows:", nrow(test_df), "\n")

# Fit ranger random forest
rf_model <- ranger(
  TRUMPSHARE ~ .,
  data       = train_df,
  num.trees  = 500,
  mtry       = 8,
  importance = "impurity",   # or "permutation" if you want stricter importance
  seed       = 42
)

# Print a brief model summary (includes OOB R²)
print(rf_model)

# Predict on the test set
rf_pred <- predict(rf_model, data = test_df)$predictions

# Evaluate performance on the test set
rss  <- sum((rf_pred - test_df$TRUMPSHARE)^2)
tss  <- sum((test_df$TRUMPSHARE - mean(test_df$TRUMPSHARE))^2)
r2   <- 1 - rss / tss
rmse <- sqrt(mean((rf_pred - test_df$TRUMPSHARE)^2))

cat("\nPerformance on test set:\n")
cat("R²   =", round(r2, 3), "\n")
cat("RMSE =", round(rmse, 4), "\n\n")

# Variable importance summary
imp <- sort(rf_model$variable.importance, decreasing = TRUE)

cat("Top 10 most important features:\n")
print(round(imp[1:10], 4))

```

0.737 R squared test set - minimal improvement with area included.


```{r}
library(sf)

# 1. Define path to the specific MI zip
mi_zip <- "data/prep/precincts/2020/raw_precinct_20/mi_2020.zip"

# 2. Unzip to a temporary folder just to look
peek_dir <- file.path(tempdir(), "mi_peek")
unzip(mi_zip, exdir = peek_dir)

# 3. Read ONLY the header/attributes (using query to keep it fast)
# Find the .shp file
mi_shp <- list.files(peek_dir, pattern = "\\.shp$", recursive = TRUE, full.names = TRUE)[1]

# Peek at names
mi_data <- st_read(mi_shp, quiet = TRUE)
print(names(mi_data))

# Optional: specifically check for long names (>10 chars) that will cause the error
long_names <- names(mi_data)[nchar(names(mi_data)) > 10]
if(length(long_names) > 0) {
  cat("\nWarning: These columns will be truncated/clash:\n")
  print(long_names)
}

# Cleanup the peek folder
unlink(peek_dir, recursive = TRUE)
```

```{r}
master <- read.csv("data/prep/precincts/2020/raw_emb_precinct_20.csv")
head(master$A00_mean)
colnames(master)

master24 <- read.csv("data/prep/precincts/2024/raw_emb_precinct_24.csv")
colnames(master24)

clean_24_emb <- read.csv("data/prep/precincts/2024/clean_emb_precinct_24.csv")
colnames(clean_24_emb)
head(clean_24_emb$state)

old <- read.csv("data/prep/precincts/2020/precinct_embeddings_2020_mean_area_std.csv")
head(old$A00_mean)

```

